---
title: "Forecasting ASX Stocks"
author: "Yemunn Soe"
date: today
format:
  html:
    code-fold: true
---

```{r setup}
#| label: setup
#| include: false
#| cache: false
# ── Setup ──────────────────────────────────────────────────────────────────────
# This chunk loads all required packages. It is never cached (cache: false)
# because packages must be loaded fresh at the start of every render session.
# If any package is missing, run: install.packages("package_name")

library(tidyquant)   # Yahoo Finance data via tq_get()
library(dplyr)       # Data wrangling: filter, mutate, arrange, group_by
library(TSA)         # Time series tools: BoxCox.ar()
library(forecast)    # auto.arima(), forecast(), checkresiduals()
library(ggplot2)     # Plotting the forecast distribution
library(purrr)       # Functional iteration: map(), map_dfr(), possibly()
library(tidyr)       # Data reshaping (used in nested tibbles)
library(magrittr)    # Pipe operator %>%
```

------------------------------------------------------------------------

# Data Acquisition {#sec-data}

## Define the ASX Top 10 Tickers

The `tidyquant` function `tq_exchange()` only supports US exchanges (AMEX, NASDAQ, NYSE) and cannot retrieve an ASX constituent list. Instead, we manually specify the **top 30 ASX stocks by market capitalisation** — the most liquid, most influential companies on the Australian market. Together they represent the core of the ASX 200's total market cap weight.

Yahoo Finance requires the `.AX` suffix for all ASX-listed stocks (e.g. `CBA` → `CBA.AX`). We store both the clean ASX symbol and the Yahoo Finance symbol for clarity.

```{r asx10-tickers}
#| label: asx10-tickers
#| cache: true

# Manually define the top 30 ASX stocks by market cap.
# Three columns are created:
#   symbol    — the official ASX ticker (used for display and filtering)
#   company   — full company name (used in tables and labels)
#   sector    — GICS sector classification (enables sector-level breakdowns later)
# A fourth column symbol_yf is derived by appending ".AX" for Yahoo Finance.

asx10_tickers <- tibble(
  symbol  = c(
    "BHP",  "CBA",  "CSL",  "NAB",  "WBC",
    "ANZ",  "WES",  "MQG",  "RIO",  "TLS"
  ),
  company = c(
    "BHP Group",               "Commonwealth Bank",       "CSL Limited",
    "National Australia Bank", "Westpac Banking",         "ANZ Banking Group",
    "Wesfarmers",              "Macquarie Group",         "Rio Tinto",
    "Telstra"
  ),
  sector  = c(
    "Materials",   "Financials",  "Healthcare",
    "Financials",  "Financials",  "Financials",
    "Consumer",    "Financials",  "Materials",
    "Telecom"
  )
) %>%
  mutate(symbol_yf = paste0(symbol, ".AX"))

cat("Tickers defined:", nrow(asx10_tickers), "\n")
print(asx10_tickers[, c("symbol", "company", "sector", "symbol_yf")])
```

## Download Daily Price History

We use `tq_get()` to pull **two years of daily** Open, High, Low, Close, and Volume **(OHLCV) data** from Yahoo Finance for each ticker. The adjusted close price — which accounts for dividends and stock splits — is the only column we use in this analysis. The dataset comprises daily trading data from April 1, 2024, to March 31, 2026 (approximately 5080 observations of 10 different stocks), extracted via Yahoo Finance using the `tidyquant` package.

The chunk is **cached**: once downloaded, data is stored in `_cache/` and Yahoo Finance is not queried again unless you change the chunk label or delete the cache folder. This prevents unnecessary re-downloads on every render.

We wrap `tq_get()` in `purrr::possibly()` so that any single ticker that fails (delisted stock, network timeout, temporary Yahoo outage) is silently skipped rather than crashing the pipeline.

But for this analysis, we typically examine: **Price and Volumes** of selected stocks

**Characteristics of Data:**

1.  **Frequency:** Daily (Monday–Friday trading days, excluding weekends and public holidays)
2.  **Sample Period:** 2024-04-01 to 2026-03-31 (2 calendar years)
3.  **Variables:**
    -   `adjusted`: Closing price adjusted for corporate actions (splits, dividends)
    -   `volume`: Number of shares traded - `date`: Trading day timestamp

```{r asx10-prices}
#| label: asx10-prices
#| cache: true

# possibly() wraps tq_get() in a try-catch:
# on failure it returns NULL instead of an error.
# map_dfr() row-binds all successful results, silently dropping NULLs.
safe_tq_get <- possibly(tq_get, otherwise = NULL)

asx10_prices_raw <- asx10_tickers$symbol_yf %>%
  map_dfr(~ safe_tq_get(
    .x,
    get  = "stock.prices",
    from = Sys.Date() - 730,   # ~2 calendar years of daily data
    to   = Sys.Date()
  ))

  
cat("Total rows downloaded:", nrow(asx10_prices_raw), "\n")
cat("Unique tickers retrieved:", n_distinct(asx10_prices_raw$symbol), "\n")
```

## Time Series Plots — Price Level & Volumes

For this we choose Commonwealth Bank of Australia (CBA), one of Australia's "Big Four" banks and a key component of the ASX 10 index, as a example. Before fitting any forecasting model, we must understand the underlying structure of our series. The first diagnostic step is always a visual inspection. Time series decomposition aims to separate a series into three fundamental components:

1.  **Trend:** Long-term directional movement (upward/downward/stationary)
2.  **Seasonality:** Regular periodic fluctuations (patterns)
3.  **Change Point**
4.  **Changing Variance**
5.  **Behavior**

```{r cba-ts-plots}
#| label: cba-ts-plots
#| fig-cap: "CBA.AX adjusted close price (top) vs. daily log returns (bottom). The price level drifts upward with no stable mean — a clearly nonstationary series. Log returns fluctuate around zero with no trend, consistent with stationarity."

# First filter for CBA stock only
cba_data <- asx10_prices_raw %>%
  filter(symbol == "CBA.AX") %>%
  arrange(date)  # Ensure chronological order

# Now create ts objects for CBA
cba_prices_ts <- ts(
  cba_data$adjusted,
  start = c(2024, 4),  # April 2024
  frequency = 365
)

cba_volumes_ts <- ts(
  cba_data$volume,
  start = c(2024, 4),
  frequency = 365
)


# Two-panel layout: price level on top, returns on bottom.
par(mfrow = c(2, 1), cex = 0.85)

plot(cba_prices_ts,
     type = "l",  # "l" for line (better than "o" for daily data)
     ylab = "Price",
     main = "CBA Daily Adjusted Price")

plot(cba_volumes_ts,
     type = "l",
     ylab = "Volumes",
     main = "CBA Daily Volumes")

par(mfrow = c(1, 1))
```

Text

```{r cba-acf-pacf-plots}
#| label: cba-acf-pacf-plots
#| fig-cap: "Diagnostic plots for CBA.AX: Autocorrelation (ACF) and Partial Autocorrelation (PACF) functions for daily adjusted closing prices and trading volumes."

# Set up a 2x2 grid for diagnostic plots
par(mfrow = c(2, 2))

# Row 1: Price Analysis
acf(cba_prices_ts, 
    main = "ACF — CBA Daily Price")

pacf(cba_prices_ts, 
     main = "PACF — CBA Daily Price")

# Row 2: Volume Analysis
acf(cba_volumes_ts, 
    main = "ACF — CBA Daily Volumes")

pacf(cba_volumes_ts, 
     main = "PACF — CBA Daily Volumes")

# Reset plotting parameters to default
par(mfrow = c(1, 1))
```

#### Transformation & Differencing

Text

```{r cba-price-transformation-plot}
#| label: cba-price-transformation-plot
#| fig-cap: "CBA.AX price transformations: (a) Original daily price; (b) Log-transformed prices stabilizing variance while preserving trend; (c) First difference of log-transformed prices achieving stationarity with fluctuations around zero."

# Set up a 2x2 grid for plots
par(mfrow = c(3, 1))

# Panel 1: Original series
plot(cba_prices_ts,
     type = "l",  
     ylab = "Price",
     main = "Original CBA Daily Price")

# Panel 2: Log transformation stabilises variance
# The dramatic spike is compressed, making growth and decay more symmetric
plot(log(cba_prices_ts),
     type = "l",  
     ylab = "log(Price)",
     main = "Log-Transformed")

# Panel 3: First difference of the log ≈ daily growth rate
# The series should now fluctuate around zero if stationarity is achieved
plot(diff(log(cba_prices_ts), differences = 1), 
     type = "l",  
     ylab = "Δ log(Price)",
     main = "1st Difference of log-Transformed")


# Reset plotting parameters to default
par(mfrow = c(1, 1))
```

#### Text

```{r cba-volume-transformation-plot}
#| label: cba-volume-transformation-plot
#| fig-cap: "CBA.AX volumes transformations: (a) Original daily volumes; (b) Log-transformed volumes stabilizing variance while preserving trend; (c) First difference of log-transformed volumes achieving stationarity with fluctuations around zero."

# Set up a 2x2 grid for plots
par(mfrow = c(3, 1))

# Panel 1: Original series
plot(cba_volumes_ts,
     type = "l",  
     ylab = "Price",
     main = "Original CBA Daily Price")

# Panel 2: Log transformation stabilises variance
# The dramatic spike is compressed, making growth and decay more symmetric
plot(log(cba_volumes_ts),
     type = "l",  
     ylab = "log(Price)",
     main = "Log-Transformed")

# Panel 3: First difference of the log ≈ daily growth rate
# The series should now fluctuate around zero if stationarity is achieved
plot(diff(log(cba_volumes_ts), differences = 1), 
     type = "l",  
     ylab = "Δ log(Price)",
     main = "1st Difference of log-Transformed")


# Reset plotting parameters to default
par(mfrow = c(1, 1))
```

#### 
