---
title: "Forecasting ASX Stocks"
subtitle: "Time Series Analysis of CBA Weekly Adjusted Closing Price"
author: "Yemunn Soe"
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: true
    fig-cap-location: bottom
    theme: flatly
---

```{r setup}
#| label: setup
#| include: false
#| cache: false

rm(list = ls())

library(tidyquant)
library(tidyverse)
library(dplyr)
library(TSA)
library(forecast)
library(tseries)
library(lmtest)
library(ggplot2)
library(purrr)
library(magrittr)
library(knitr)
library(kableExtra)
```

```{r functions}
#| label: functions
#| include: false

# ══════════════════════════════════════════════════════════════════════════════
# HELPER FUNCTIONS
# A. Plotting helpers     (1–3, 8, 20–21)
# B. Formal test helpers  (4–6)
# C. Residual diagnostics (9–10)
# D. Output helpers       (7)
# E. Forecasting          (12–19)
# ══════════════════════════════════════════════════════════════════════════════

# ── Function 1: Time series plot ──────────────────────────────────────────────
plot_ts <- function(ts_data, main, ylab = "Price", add_mean = FALSE) {
  plot(ts_data, type = "o", ylab = ylab, xlab = "Time",
       cex.main = 0.9, main = main)
}

# ── Function 2: Side-by-side ACF and PACF plots ───────────────────────────────
plot_acf_pacf <- function(ts_data, acf_title, pacf_title, lag.max = 60) {
  par(mfrow = c(1, 2), cex.main = 0.9)
  acf(ts_data,  lag.max = lag.max, main = acf_title,  ylab = "ACF",  xlab = "Lag")
  pacf(ts_data, lag.max = lag.max, main = pacf_title, ylab = "PACF", xlab = "Lag")
  par(mfrow = c(1, 1))
}

# ── Function 3: Side-by-side QQ plot and Histogram ────────────────────────────
plot_norm <- function(ts_data, qq_title, hist_title, xlab = "Value") {
  par(mfrow = c(1, 2), cex.main = 0.9)
  qqnorm(ts_data, main = qq_title, pch = 16, cex = 0.7)
  qqline(ts_data, col = "blue", lty = 2, lwd = 1.5)
  hist(ts_data, xlab = xlab, main = hist_title,
       border = "black", breaks = 20)
  par(mfrow = c(1, 1))
}

# ── Function 4: Shapiro-Wilk normality test with kable output ─────────────────
check_normality <- function(ts_data, caption_text) {
  sw <- shapiro.test(as.numeric(ts_data))
  p_formatted <- format(sw$p.value, scientific = TRUE, digits = 2)
  df <- data.frame(
    "Test Statistic (W)" = round(sw$statistic, 4),
    "p-value"            = p_formatted,
    check.names          = FALSE
  )
  list(
    result = sw,
    table  = knitr::kable(df, caption = caption_text,
                          align = "c", row.names = FALSE)
  )
}

# ── Function 5: Single ADF test with kable output ─────────────────────────────
check_adf <- function(ts_data, caption_text) {
  adf_res     <- adf.test(ts_data, alternative = "stationary")
  p_formatted <- format(adf_res$p.value, scientific = TRUE, digits = 2)
  stat_value  <- as.numeric(adf_res$statistic)
  df <- data.frame(
    "Test Statistic" = round(stat_value, 4),
    "p-value"        = p_formatted,
    check.names      = FALSE
  )
  list(
    result = adf_res,
    table  = knitr::kable(df, caption = caption_text,
                          align = "c", row.names = FALSE)
  )
}

# ── Function 6: Full stationarity battery (ADF + PP + KPSS) ──────────────────
check_stationarity <- function(ts_data, table_title) {
  adf_res  <- adf.test(ts_data,  alternative = "stationary")
  pp_res   <- pp.test(ts_data,   alternative = "stationary")
  kpss_res <- kpss.test(ts_data, null        = "Trend")

  adf_v  <- ifelse(adf_res$p.value  < 0.05,
                   "Reject H0 → Stationary",
                   "Fail to Reject H0 → Non-stationary")
  pp_v   <- ifelse(pp_res$p.value   < 0.05,
                   "Reject H0 → Stationary",
                   "Fail to Reject H0 → Non-stationary")
  kpss_v <- ifelse(kpss_res$p.value > 0.05,
                   "Fail to Reject H0 → Stationary",
                   "Reject H0 → Non-stationary")

  df <- data.frame(
    "Test"      = c("Augmented Dickey-Fuller (ADF)",
                    "Phillips-Perron (PP)",
                    "KPSS"),
    "H0"        = c("Unit root (non-stationary)",
                    "Unit root (non-stationary)",
                    "Series is stationary"),
    "Statistic" = c(round(unname(adf_res$statistic),  4),
                    round(unname(pp_res$statistic),   4),
                    round(unname(kpss_res$statistic), 4)),
    "p-value"   = c(format(adf_res$p.value,  scientific = TRUE, digits = 2),
                    format(pp_res$p.value,   scientific = TRUE, digits = 2),
                    format(kpss_res$p.value, scientific = TRUE, digits = 2)),
    "Decision"  = c(adf_v, pp_v, kpss_v),
    check.names = FALSE
  )
  list(
    adf   = adf_res,
    pp    = pp_res,
    kpss  = kpss_res,
    table = knitr::kable(df, caption = table_title, align = "llccc")
  )
}

# ── Function 7: Coefficient significance table ────────────────────────────────
make_coef_table <- function(model, caption_text) {
  ct <- lmtest::coeftest(model)
  df <- data.frame(
    Coefficient = rownames(ct),
    Estimate    = round(ct[, 1], 5),
    Std.Error   = round(ct[, 2], 5),
    z.value     = round(ct[, 3], 4),
    p.value     = round(ct[, 4], 4),
    Sig         = ifelse(ct[, 4] < 0.001, "***",
                  ifelse(ct[, 4] < 0.01,  "**",
                  ifelse(ct[, 4] < 0.05,  "*",
                  ifelse(ct[, 4] < 0.1,   ".", ""))))
  )
  knitr::kable(df, caption = caption_text, align = "c", row.names = FALSE)
}

# ── Function 8: ML fitted value overlay plot ──────────────────────────────────
plot_ml_fitted <- function(ml_model, ts_data, spec) {
  par(mar = c(4, 4, 3, 2))
  obs_col <- adjustcolor("black", alpha.f = 0.85)
  plot(ts_data, type = "o", ylab = "Price (AUD)", xlab = "Time",
       main = paste(spec, "(ML)"), col = obs_col, pch = 16, cex = 0.5)
  lines(fitted(ml_model), col = "blue", lwd = 2)
  legend("topleft", legend = c("Observed", "Fitted"),
         col = c(obs_col, "blue"), lty = 1, lwd = c(1, 2),
         cex = 0.65, bty = "n")
  par(mfrow = c(1, 1))
}

# ── Function 9: Residual visual panel (3×2 grid) ──────────────────────────────
plot_residual_diagnostics <- function(model, model_name, lag.max = 40) {
  res     <- residuals(model)
  std_res <- rstandard(model)
  lb_pvalues <- numeric(lag.max)
  for (i in 1:lag.max) {
    lb_pvalues[i] <- Box.test(res, lag = i, type = "Ljung-Box")$p.value
  }
  par(mfrow = c(3, 2), mar = c(4, 4, 3, 2))
  # Panel 1: Residuals over time
  plot(res, type = "l", xlab = "Time", ylab = "Residuals",
       main = paste(model_name, "— Residuals over Time"), col = "black")
  points(res, pch = 16, cex = 0.5, col = "black")
  abline(h = 0, col = "blue", lty = 2)
  # Panel 2: Standardised residuals
  plot(std_res, type = "l", xlab = "Time", ylab = "Standardised Residuals",
       main = paste(model_name, "— Standardised Residuals"), col = "black")
  points(std_res, pch = 16, cex = 0.5, col = "black")
  abline(h =  0, col = "blue", lty = 2)
  abline(h =  3, col = "red",  lty = 3)
  abline(h = -3, col = "red",  lty = 3)
  # Panel 3: Histogram
  hist(res, breaks = 20, xlab = "Residuals",
       main = paste(model_name, "— Histogram"),
       border = "black", freq = TRUE)
  # Panel 4: QQ plot
  qqnorm(res, main = paste(model_name, "— QQ Plot"), pch = 16, cex = 0.7)
  qqline(res, col = "blue", lty = 2, lwd = 1.5)
  # Panel 5: ACF of residuals
  acf(res, lag.max = lag.max,
      main = paste(model_name, "— ACF of Residuals"),
      ylab = "ACF", xlab = "Lag")
  # Panel 6: Ljung-Box p-values
  plot(1:lag.max, lb_pvalues, type = "p", pch = 1, col = "black",
       xlab = "Lag", ylab = "p-value",
       main = "Ljung-Box p-values", ylim = c(0, 1))
  abline(h = 0.05, col = "blue", lty = 2, lwd = 1.5)
  par(mfrow = c(1, 1))
  invisible(list(residuals = res, std_res = std_res, lb_pvalues = lb_pvalues))
}

# ── Function 10: Formal residual tests (Ljung-Box + Shapiro-Wilk) ─────────────
test_residuals <- function(model, model_name, lag.max = 40) {
  res <- rstandard(model)
  lb  <- Box.test(res, lag = min(lag.max, floor(length(res) / 5)),
                  type = "Ljung-Box")
  sw  <- shapiro.test(as.numeric(res))
  df  <- data.frame(
    "Test"      = c("Ljung-Box", "Shapiro-Wilk"),
    "H0"        = c("No autocorrelation in residuals",
                    "Residuals are normally distributed"),
    "Statistic" = c(round(lb$statistic, 4), round(sw$statistic, 4)),
    "df"        = c(lb$parameter, NA),
    "p-value"   = c(round(lb$p.value, 4), round(sw$p.value, 4)),
    "Decision"  = c(ifelse(lb$p.value > 0.05,
                           "Fail to Reject H0 → PASS", "Reject H0 → FAIL"),
                    ifelse(sw$p.value > 0.05,
                           "Fail to Reject H0 → PASS", "Reject H0 → FAIL")),
    check.names = FALSE
  )
  list(lb = lb, sw = sw,
       table = knitr::kable(df,
                            caption = paste("Formal residual tests —", model_name),
                            align   = "llcccc"))
}

# ── Function 12: Benchmark forecasts (Naive, Drift, Mean) ─────────────────────
benchmark_forecasts <- function(y_train, y_test) {
  h <- length(y_test)
  list(
    Naive = forecast(naive(y_train, h = h)),
    Drift = rwf(y_train, drift = TRUE, h = h),
    Mean  = forecast(meanf(y_train, h = h))
  )
}

# ── Function 13: ETS forecast ─────────────────────────────────────────────────
ets_forecast <- function(y_train, h) {
  m <- ets(y_train)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 14: auto.arima forecast ──────────────────────────────────────────
auto_arima_forecast <- function(y_train, h, seasonal = FALSE) {
  m <- auto.arima(y_train, seasonal = seasonal,
                  stepwise = FALSE, approximation = FALSE)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 15: TBATS forecast ───────────────────────────────────────────────
tbats_forecast <- function(y_train, h) {
  m <- tbats(y_train)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 16: NNETAR forecast ──────────────────────────────────────────────
nnetar_forecast <- function(y_train, h, P = 1, repeats = 20) {
  m <- nnetar(y_train, P = P, repeats = repeats)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 17: STL decomposition forecast ───────────────────────────────────
stl_forecast <- function(y_train, h, s.window = 7) {
  m <- stlm(y_train, modelfunction = ar, s.window = s.window)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 18: Forecast comparison table ────────────────────────────────────
compare_forecasts <- function(forecast_list, y_test,
                              caption = "Forecast Accuracy Comparison") {
  acc_df <- data.frame()
  for (name in names(forecast_list)) {
    fc_obj <- forecast_list[[name]]
    if (!is.null(fc_obj) && !is.null(fc_obj$mean)) {
      acc <- tryCatch(accuracy(fc_obj, y_test), error = function(e) NULL)
      if (is.null(acc)) next
      if (is.matrix(acc) && nrow(acc) >= 2) acc <- acc[2, ]
      temp_df <- data.frame(
        Method = name,
        RMSE   = round(as.numeric(acc["RMSE"]), 4),
        MAE    = round(as.numeric(acc["MAE"]),  4),
        MAPE   = round(as.numeric(acc["MAPE"]), 4),
        MASE   = round(as.numeric(acc["MASE"]), 4),
        stringsAsFactors = FALSE
      )
      acc_df <- rbind(acc_df, temp_df)
    }
  }
  acc_df      <- acc_df[order(acc_df$RMSE), ]
  acc_df$Rank <- seq_len(nrow(acc_df))
  acc_df      <- acc_df[, c("Rank", "Method", "RMSE", "MAE", "MAPE", "MASE")]
  knitr::kable(acc_df, caption = caption, align = "clcccc",
               row.names = FALSE) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}

# ── Function 19: Ensemble forecast ────────────────────────────────────────────
ensemble_forecast <- function(forecast_list, weights = NULL) {
  valid     <- Filter(function(x) !is.null(x) && !is.null(x$mean), forecast_list)
  fc_matrix <- do.call(cbind, lapply(valid, function(x) as.numeric(x$mean)))
  if (is.null(weights)) weights <- rep(1 / ncol(fc_matrix), ncol(fc_matrix))
  ens_mean  <- as.numeric(fc_matrix %*% weights)
  ts(ens_mean, start = start(valid[[1]]$mean), frequency = frequency(valid[[1]]$mean))
}

# ── Function 20: Forecast plot (single model with PIs) ────────────────────────
plot_forecast_simple <- function(fc, y_train, y_test, model_name) {
  h     <- length(y_test)
  x_all <- c(as.numeric(time(y_train)), as.numeric(time(y_test)))
  y_all <- c(as.numeric(y_train),       as.numeric(y_test))
  plot(x_all, y_all, type = "n",
       main = paste(model_name, "— Forecast, Next", h, "Weeks"),
       ylab = "Price (AUD)", xlab = "Time")
  if (!is.null(fc$lower) && !is.null(fc$upper)) {
    fc_time <- as.numeric(time(fc$mean))
    polygon(c(fc_time, rev(fc_time)),
            c(as.numeric(fc$lower[, "95%"]),
              rev(as.numeric(fc$upper[, "95%"]))),
            col = adjustcolor("steelblue", alpha.f = 0.2), border = NA)
    polygon(c(fc_time, rev(fc_time)),
            c(as.numeric(fc$lower[, "80%"]),
              rev(as.numeric(fc$upper[, "80%"]))),
            col = adjustcolor("steelblue", alpha.f = 0.4), border = NA)
  }
  lines(as.numeric(time(y_train)), as.numeric(y_train),
        col = "black",      lwd = 1.5)
  lines(as.numeric(time(fc$mean)), as.numeric(fc$mean),
        col = "steelblue4", lwd = 2)
  lines(as.numeric(time(y_test)),  as.numeric(y_test),
        col = "red",        lwd = 2)
  legend("topleft",
         legend = c("Training", "Forecast", "80% PI", "95% PI", "Actual"),
         col    = c("black", "steelblue4",
                    adjustcolor("steelblue", 0.4),
                    adjustcolor("steelblue", 0.2), "red"),
         lwd = c(1.5, 2, 8, 8, 2), bty = "n", cex = 0.7)
}

# ── Function 21: Multi-method forecast comparison plot ────────────────────────
plot_forecast_comparison <- function(forecast_list, y_train, y_test,
                                     title = "Forecast Comparison") {
  all_fc_vals <- unlist(lapply(forecast_list,
                               function(x) if (!is.null(x$mean)) as.numeric(x$mean)))
  ylim_range  <- range(c(as.numeric(y_train), as.numeric(y_test),
                          all_fc_vals), na.rm = TRUE)
  x_all <- c(as.numeric(time(y_train)), as.numeric(time(y_test)))
  y_all <- c(as.numeric(y_train),       as.numeric(y_test))
  plot(x_all, y_all, type = "n",
       xlab = "Time", ylab = "Price (AUD)", main = title, ylim = ylim_range)
  grid(col = "lightgray")
  lines(as.numeric(time(y_train)), as.numeric(y_train), col = "black", lwd = 2)
  lines(as.numeric(time(y_test)),  as.numeric(y_test),  col = "red",   lwd = 2, lty = 2)
  pal <- c("blue", "darkgreen", "purple", "darkorange",
           "brown", "deeppink", "cyan4", "navy", "tomato")
  for (i in seq_along(forecast_list)) {
    fc <- forecast_list[[i]]
    if (!is.null(fc) && !is.null(fc$mean)) {
      lines(as.numeric(time(fc$mean)), as.numeric(fc$mean),
            col = pal[(i - 1) %% length(pal) + 1], lwd = 1.5, lty = 1)
    }
  }
  legend("topleft",
         legend = c("Training", "Actual (test)", names(forecast_list)),
         col    = c("black", "red", pal[seq_along(forecast_list)]),
         lty    = c(1, 2, rep(1, length(forecast_list))),
         lwd = 2, cex = 0.6, bty = "n")
}
```

## Introduction

This study applies a systematic ARIMA time series modelling workflow to the weekly adjusted closing price of Commonwealth Bank of Australia (CBA.AX) — one of Australia's "Big Four" banks and the largest company by market capitalisation on the ASX 200. CBA's price movements are closely watched as a barometer of the broader Australian financial sector, domestic housing market conditions, and the Reserve Bank of Australia's (RBA) monetary policy trajectory.

The analysis follows:

-   Exploratory Data Analysis — visual inspection of trend, variance, and distributional properties.
-   Transformation — Box-Cox variance stabilisation where warranted.
-   Differencing & Stationarity — achieving a time-invariant mean via differencing, verified by formal unit-root tests.
-   Model Specification — AR and MA order identification via ACF, PACF, EACF, and BIC tables.
-   Model Fitting & Residual Diagnostics — CSS and ML estimation with full residual analysis and overparameterised model checks.
-   Forecasting — producing point forecasts and prediction intervals using multiple methods.

## Data Acquisition

### Define ASX Top 10 Tickers

```{r asx10-tickers}
#| label: asx10-tickers
#| cache: true

asx10_tickers <- tibble(
  symbol  = c("BHP",  "CBA",  "CSL",  "NAB",  "WBC",
               "ANZ",  "WES",  "MQG",  "RIO",  "TLS"),
  company = c("BHP Group",               "Commonwealth Bank",
               "CSL Limited",             "National Australia Bank",
               "Westpac Banking",         "ANZ Banking Group",
               "Wesfarmers",              "Macquarie Group",
               "Rio Tinto",               "Telstra"),
  sector  = c("Materials",   "Financials",  "Healthcare",
               "Financials",  "Financials",  "Financials",
               "Consumer",    "Financials",  "Materials",  "Telecom")
) %>%
  mutate(symbol_yf = paste0(symbol, ".AX"))

cat("Tickers defined:", nrow(asx10_tickers), "\n")
print(asx10_tickers[, c("symbol", "company", "sector", "symbol_yf")])
```

### Download Weekly Price History

```{r asx10-prices}
#| label: asx10-prices
#| cache: true

safe_tq_get <- possibly(tq_get, otherwise = NULL)

asx10_prices_raw <- asx10_tickers$symbol_yf %>%
  map_dfr(~ safe_tq_get(
    .x,
    get         = "stock.prices",
    from        = Sys.Date() - (5 * 365),
    to          = Sys.Date(),
    periodicity = "weekly"
  ))

cat("Total rows downloaded:", nrow(asx10_prices_raw), "\n")
cat("Unique tickers retrieved:", n_distinct(asx10_prices_raw$symbol), "\n")
```

### Construct Time Series Objects

```{r cba-ts-setup}
#| label: cba-ts-setup
#| cache: true

cba_data <- asx10_prices_raw %>%
  filter(symbol == "CBA.AX") %>%
  arrange(date)

cba_prices_ts <- ts(
  cba_data$adjusted,
  start     = c(year(min(cba_data$date)), week(min(cba_data$date))),
  frequency = 52
)

n_train <- length(cba_prices_ts) - 12   # 12 weeks = 1 quarter held out
y_train <- window(cba_prices_ts, end   = time(cba_prices_ts)[n_train])
y_test  <- window(cba_prices_ts, start = time(cba_prices_ts)[n_train + 1])

cat("CBA observations (total):", length(cba_prices_ts), "\n")
cat("Training observations:   ", length(y_train), "\n")
cat("Test observations:       ", length(y_test),  "\n")
cat("Date range:", format(min(cba_data$date), "%d %b %Y"),
    "to",         format(max(cba_data$date), "%d %b %Y"), "\n")
cat("Price range: AUD", round(min(cba_data$adjusted), 2),
    "–",               round(max(cba_data$adjusted), 2), "\n")
```

## Exploratory Data Analysis (EDA)

### Stationarity Inspection

#### Visual Inspection

```{r fig-eda-raw-ts}
#| label: fig-eda-raw-ts
#| fig-cap: "CBA adjusted closing price — training sample. The plot visualises the stock's historical performance from late 2021 through early 2026."
#| fig-height: 4

plot_ts(y_train, main = "CBA Adjusted Closing Price — Training Sample",
        ylab = "Price (AUD)", add_mean = TRUE)
```

```{r fig-eda-acf-pacf-raw}
#| label: fig-eda-acf-pacf-raw
#| fig-cap: "ACF (left) and PACF (right) of the raw CBA price series. The ACF's slow linear decay and the PACF's single dominant spike at lag 1 confirm non-stationarity before any statistical testing."
#| fig-height: 4

plot_acf_pacf(y_train,
              acf_title  = "ACF — CBA Raw Price",
              pacf_title = "PACF — CBA Raw Price")
```

#### Statistical Tests

```{r stat-eda-stationarity}
#| label: stat-eda-stationarity

stat_raw <- check_adf(y_train,
  "Stationarity Test — Raw CBA Price (Training Sample)")
stat_raw$table
```

#### Interpretation

**Trend.** @fig-eda-raw-ts shows a sustained upward quadratic trend from approximately AUD `r round(min(y_train), 2)` at the start to a peak near AUD `r round(max(y_train), 2)` by mid-2025. This sustained growth indicates a non-stationary mean, confirming that the series requires mean stabilisation before modelling.

**Seasonality.** There is no evidence of a fixed-frequency repeating cycle within this observed window. Price fluctuations appear stochastic rather than deterministic. Any apparent periodicity aligns with irregular macroeconomic events rather than a strict calendar pattern. This suggests that a non-seasonal ARIMA specification is appropriate, and no seasonal differencing or seasonal parameters are required.

**Heteroscedasticity.** The variance appears relatively stable (homoscedastic) throughout the period. The magnitude of fluctuations does not increase proportionally with the mean, suggesting the series may not require a variance-stabilising transformation.

**Behavior.** The series demonstrates strong autoregressive (AR) behavior, where each observation is highly dependent on its previous value. The sustained climb suggests that current resistance levels serve as a strong predictor for the following period.

**Change point.** No change point is observed.

::: callout-note
A **change point** in time series analysis refers to a transition where the underlying statistical properties of the data — such as the mean, variance, or trend — undergo a sudden and persistent shift.
:::

**ACF — slow linear decay** (@fig-eda-acf-pacf-raw, left). The autocorrelations remain large and positive across many lags, declining so slowly that they barely enter the significance bands even at lag 60. This persistence confirms the series has very long memory: each week's adjusted price level carries the weight of all preceding observations forward indefinitely.

**PACF — single dominant spike at lag 1** (@fig-eda-acf-pacf-raw, right). Once the chain of correlations is removed, only the immediately preceding observation retains direct predictive power. This is the canonical PACF signature of a unit-root process.

**Stationarity tests.** The series contains a stochastic trend. The ADF (p = `r round(stat_raw$result$p.value, 4)`) fails to reject the unit-root null, confirming the non-stationarity that must be removed before modelling.

### Normality & Distributional Inspection

#### Visual Inspection

```{r fig-eda-norm}
#| label: fig-eda-norm
#| fig-cap: "QQ plot (left) and histogram (right) for the raw CBA training price series. The S-shaped deviation from the reference line and the non-symmetric histogram reject the assumption that the data follow a normal distribution."
#| fig-height: 4

plot_norm(y_train, qq_title  = "QQ Plot — Raw CBA Price",
          hist_title = "Histogram — Raw CBA Price", xlab = "Price (AUD)")
```

#### Statistical Tests

```{r stat-eda-norm-summary}
#| label: stat-eda-norm-summary

Summary <- c(
  Mean   = round(mean(y_train),               2),
  Median = round(median(y_train),             2),
  SD     = round(sd(y_train),                 2),
  Min    = round(min(y_train),                2),
  Max    = round(max(y_train),                2),
  Range  = round(max(y_train) - min(y_train), 2)
)
knitr::kable(t(as.data.frame(Summary)),
  caption = "Descriptive statistics — CBA.AX adjusted closing price (training sample).",
  align   = "c")
```

```{r stat-eda-normality}
#| label: stat-eda-normality

normality_raw <- check_normality(y_train, "Shapiro-Wilk Test — Raw CBA Price")
normality_raw$table
```

#### Interpretation

**Central tendency.** The mean (AUD `r round(mean(y_train), 2)`) and median (AUD `r round(median(y_train), 2)`) are closely aligned, but this proximity is misleading: it results from the series spending roughly equal time in its lower and upper price ranges, not from any genuine normality.

**Dispersion.** The standard deviation of AUD `r round(sd(y_train), 2)` reflects the wide price range traversed over the training window — a direct consequence of the sustained upward trend in @fig-eda-raw-ts rather than a meaningful measure of single-period volatility.

**QQ plot** (@fig-eda-norm, left). The S-shaped deviation from the theoretical line at both tails indicates lighter tails than a normal distribution — a natural consequence of the bounded lower tail (prices cannot fall below zero) and the right-skew imposed by the trend.

**Histogram** (@fig-eda-norm, right). The multimodal profile with at least two visible clusters corresponds to the lower pre-2025 and higher 2025-peak price ranges. No single normal density can describe this bimodal structure, which is a direct symptom of the non-stationarity identified above.

**Shapiro-Wilk test.** With W = `r round(normality_raw$result$statistic, 4)` and p ≈ `r format(normality_raw$result$p.value, scientific = TRUE, digits = 2)`, we decisively reject normality. Transformation and differencing are required before valid ARIMA estimation.

### Synthesis of EDA Findings

**Variance stabilisation required.** The Shapiro-Wilk test (p ≈ 0), S-shaped QQ deviations, and visually apparent heteroscedasticity in @fig-eda-raw-ts collectively confirm non-constant variance proportional to price level. A Box-Cox transformation will be evaluated in the next section.

**Mean stabilisation required.** The ADF (p = `r round(stat_raw$result$p.value, 4)`) confirms the presence of a unit root. The slow ACF decay in @fig-eda-acf-pacf-raw provides visual corroboration. First-order differencing (d = 1) will be applied to address this.

## Transformation {#sec-transformation}

### Box-Cox Profile Analysis

#### Visual Inspection

```{r fig-boxcox-profile}
#| label: fig-boxcox-profile
#| fig-cap: "The Box-Cox log-likelihood profile for the CBA training series. The peak of the profile indicates the optimal λ, and the dashed lines bound the 95% confidence interval."
#| fig-height: 4

BC <- BoxCox.ar(y = y_train, lambda = seq(-2, 2, 0.01))
```

#### Statistical Tests

```{r stat-boxcox-ci}
#| label: stat-boxcox-ci

lambda_opt   <- BC$lambda[which.max(BC$loglike)]
lambda_lower <- BC$ci[1]
lambda_upper <- BC$ci[2]

bc_df <- data.frame(
  Parameter = c("Optimal λ", "95% CI Lower", "95% CI Upper",
                "CI includes λ = 0 (log)?", "CI includes λ = 1 (none)?"),
  Value     = c(round(lambda_opt,   4),
                round(lambda_lower, 4),
                round(lambda_upper, 4),
                ifelse(lambda_lower <= 0 & lambda_upper >= 0, "YES", "NO"),
                ifelse(lambda_lower <= 1 & lambda_upper >= 1, "YES", "NO"))
)
knitr::kable(bc_df, caption = "Box-Cox parameter estimation results.",
             align = "lc", col.names = c("Parameter", "Value"))
```

#### Interpretation

The optimal λ (`r round(lambda_opt, 4)`) sits close to zero, and the 95% CI from @fig-boxcox-profile determines whether a log transformation is statistically supported. For CBA weekly prices, the CI typically includes λ = 0 and excludes λ = 1 — providing formal support for the log transform and aligning with financial econometric theory: log-prices follow a random walk with drift, converting multiplicative price movements into additive log-returns.

### Applied Transformation

#### Visual Inspection

```{r fig-bc-compare}
#| label: fig-bc-compare
#| fig-cap: "Original CBA price series (top) and Box-Cox transformed series (bottom). A successful transformation would show reduced heteroscedasticity — specifically, fluctuation amplitude becoming more uniform across the sample."
#| fig-height: 8

BC_train <- (y_train^lambda_opt - 1) / lambda_opt
BC_test  <- (y_test^lambda_opt  - 1) / lambda_opt

par(mfrow = c(2, 1))
plot_ts(y_train,  main = "CBA — Original Series",            ylab = "Price (AUD)")
plot_ts(BC_train, main = "CBA — Box-Cox Transformed Series", ylab = "BC(Price)")
par(mfrow = c(1, 1))
```

```{r fig-bc-acf-pacf}
#| label: fig-bc-acf-pacf
#| fig-cap: "ACF (left) and PACF (right) of the Box-Cox transformed series. The slow ACF decay and single dominant PACF spike at lag 1 mirror the patterns in @fig-eda-acf-pacf-raw, confirming the transformation has not altered the fundamental non-stationary structure."
#| fig-height: 4

plot_acf_pacf(BC_train,
              acf_title  = "ACF — Box-Cox Transformed Series",
              pacf_title = "PACF — Box-Cox Transformed Series")
```

```{r fig-bc-norm}
#| label: fig-bc-norm
#| fig-cap: "QQ plot (left) and histogram (right) for the Box-Cox transformed series. Near-identical tail behaviour to the original series (@fig-eda-norm) indicates that the transformation has not materially improved distributional normality."
#| fig-height: 4

plot_norm(BC_train, qq_title   = "QQ Plot — BC Transformed Series",
          hist_title = "Histogram — BC Transformed Series", xlab = "BC(Price)")
```

#### Statistical Tests

```{r stat-bc-normality}
#| label: stat-bc-normality

normality_bc <- check_normality(BC_train,
  "Shapiro-Wilk Test — Box-Cox Transformed Series")
normality_bc$table
```

```{r stat-bc-stationarity}
#| label: stat-bc-stationarity

stat_bc <- check_adf(BC_train, "Stationarity Test — Box-Cox Transformed Series")
stat_bc$table
```

#### Interpretation

**Time series plot** (@fig-bc-compare). The Box-Cox transformed series retains the same overall shape — the upward trend persists and the amplitude of fluctuations continues to grow after 2025. The transformation has not materially altered the non-stationary structure or visibly reduced heteroscedasticity, consistent with financial series where variance is driven primarily by the stochastic trend.

**ACF/PACF** (@fig-bc-acf-pacf). The slow linear ACF decay and single dominant PACF spike at lag 1 are nearly identical to the raw series in @fig-eda-acf-pacf-raw. Non-stationarity in the mean remains unresolved.

**QQ plot and histogram** (@fig-bc-norm). The S-shaped QQ deviation and multimodal histogram are essentially indistinguishable from @fig-eda-norm. No meaningful distributional improvement has been achieved.

**Normality.** W = `r round(normality_bc$result$statistic, 4)`, p = `r format(normality_bc$result$p.value, scientific = TRUE, digits = 2)` — marginally worse than the original series. The normality null is again decisively rejected.

**Stationarity.** ADF (p = `r round(stat_bc$result$p.value, 4)`) confirms the transformed series remains non-stationary.

### Decision: Proceed Without Transformation

The table below summarises the key test statistics before and after transformation, confirming that the Box-Cox transformation offers no material improvement on either dimension.

```{r stat-bc-comparison}
#| label: stat-bc-comparison

bc_compare_df <- data.frame(
  Series        = c("Original", "Box-Cox Transformed"),
  "SW Statistic (W)" = c(round(normality_raw$result$statistic, 4),
                          round(normality_bc$result$statistic,  4)),
  "SW p-value"       = c(format(normality_raw$result$p.value, scientific = TRUE, digits = 2),
                          format(normality_bc$result$p.value,  scientific = TRUE, digits = 2)),
  "ADF Statistic"    = c(round(as.numeric(stat_raw$result$statistic), 4),
                          round(as.numeric(stat_bc$result$statistic),  4)),
  "ADF p-value"      = c(format(stat_raw$result$p.value, scientific = TRUE, digits = 2),
                          format(stat_bc$result$p.value, scientific = TRUE, digits = 2)),
  check.names = FALSE
)
knitr::kable(bc_compare_df,
  caption = "Comparison of normality and stationarity — original vs Box-Cox transformed series.",
  align   = "lcccc")
```

The Box-Cox transformation produces no substantive improvement in normality and negligible improvement in variance stability or stationarity. In accordance with the principle of parsimony (Cryer & Chan, 2008), the transformation is rejected. The analysis proceeds on the original adjusted price series. The next phase applies first-order differencing (d = 1) to stabilise the mean.

## Addressing Non-Stationarity {#sec-stationarity}

### First Differencing

#### Visual Inspection

```{r fig-diff-series}
#| label: fig-diff-series
#| fig-cap: "Original CBA price series (top) and first-differenced series (bottom). The differenced series fluctuates around zero, confirming the stochastic trend has been successfully removed."
#| fig-height: 8

diff_y_train <- diff(y_train, differences = 1)

par(mfrow = c(2, 1))
plot_ts(y_train,      main = "(a) Original CBA Price",              ylab = "Price (AUD)")
plot_ts(diff_y_train, main = "(b) First-Differenced Series (d = 1)", ylab = "Δ Price (AUD)",
        add_mean = TRUE)
par(mfrow = c(1, 1))
```

```{r fig-diff-acf-pacf}
#| label: fig-diff-acf-pacf
#| fig-cap: "ACF (left) and PACF (right) of the first-differenced CBA series. The immediate drop-off with most bars within the 95% confidence bands confirms stationarity and reveals the underlying ARMA structure for model specification."
#| fig-height: 4

plot_acf_pacf(diff_y_train,
              acf_title  = "ACF — Δ CBA Price (d = 1)",
              pacf_title = "PACF — Δ CBA Price (d = 1)")
```

#### Statistical Tests

```{r stat-diff-stationarity}
#| label: stat-diff-stationarity

stat_d1 <- check_stationarity(diff_y_train,
  "Stationarity Tests — First-Differenced Series (d = 1)")
stat_d1$table
```

#### Interpretation

**Trend removal** (@fig-diff-series). The upward trend in the original series (a) is completely removed in the first-differenced series (b). The differenced series fluctuates around a stable horizontal mean of approximately zero, confirming that first-order differencing was sufficient; a second difference is not required and would risk over-differencing.

**Volatility and outliers.** @fig-diff-series highlights elevated volatility near 2026. While the ARIMA framework models the linear autocorrelative structure, the presence of volatility clustering suggests that a GARCH extension may be necessary to model the conditional variance if ARCH effects are detected in the residuals.

**ACF & PACF** (@fig-diff-acf-pacf). Visual inspection confirms that first-order differencing (d = 1) has successfully achieved stationarity. Both plots display an immediate collapse with the majority of bars within the 95% confidence bounds. However, significant spikes at early lags suggest a residual short-term autocorrelative structure. These spikes will guide the selection of the AR (p) and MA (q) orders during model specification.

**Stationarity tests.** ADF (p = `r round(stat_d1$adf$p.value, 4)`) and PP (p = `r round(stat_d1$pp$p.value, 4)`) both reject the unit-root null; KPSS (p = `r round(stat_d1$kpss$p.value, 4)`) fails to reject its stationarity null. All three tests unanimously confirm stationarity. The integration order is confirmed as d = 1.

### Decision: Proceed With First-Differenced Original Series

Based on the evidence above, the analysis proceeds with the original (untransformed) series at d = 1. Further differencing (d = 2) is rejected for three reasons:

**Risk of over-differencing.** d = 1 has already achieved statistically confirmed stationarity across all three tests. A second difference introduces artificial moving-average structure and inflates residual variance unnecessarily.

**Loss of interpretability.** d = 1 transforms prices into weekly changes (returns) — a natural, meaningful financial quantity. d = 2 would model the "change in changes" (acceleration), which has no natural financial interpretation and is highly sensitive to noise.

**Parsimony.** A stationary d = 1 series allows for a simpler ARIMA structure less likely to overfit, preserving out-of-sample generalisation on the 12-week test set.

## Model Specification

### Order Identification from ACF and PACF

Visual inspection of the ACF and PACF plots (@fig-diff-acf-pacf) of the first-differenced series confirms stationarity, as evidenced by the rapid decay of autocorrelation coefficients toward zero. Despite this stability, isolated spikes suggest a residual autocorrelative signal. To investigate whether these early-lag dependencies are structural components or merely stochastic noise, we applied standard rules (Cryer & Chan, 2008) for translating ACF and PACF plots into ARMA orders:

-   ACF cuts off at lag q, PACF decays geometrically → MA(q)
-   PACF cuts off at lag p, ACF decays geometrically → AR(p)
-   Both decay geometrically → mixed ARMA(p, q)

From @fig-diff-acf-pacf, the patterns suggest candidates including ARIMA(4,1,2), ARIMA(4,1,3), and ARIMA(3,1,3).

### Extended Autocorrelation Function (EACF)

```{r stat-eacf}
#| label: stat-eacf

cat("=== EACF — First-Differenced CBA Price ===\n")
eacf(diff_y_train, ar.max = 10, ma.max = 10)
```

The EACF table identifies the upper-left vertex of the 'o' triangle, which represents candidates that effectively reduce the series to white noise. The vertex at AR = 0, MA = 6 identifies ARIMA(0,1,6) as the most parsimonious candidate, while a secondary 'o' at (1,6) suggests ARIMA(1,1,6) as a viable mixed alternative. ARIMA(0,1,7) is also included to ensure robust coverage.

### BIC Subset Selection

#### 10×10 Search Space

```{r fig-bic-10x10-subsets}
#| label: fig-bic-10x10-subsets
#| fig-cap: "BIC subset selection — 10×10 search space. The top-ranked row (lowest BIC) identifies the most parsimonious model across a broad AR/MA order grid. Shaded cells show retained parameters at each order level."
#| fig-height: 5

res_bic_10 <- armasubsets(y = diff_y_train, nar = 10, nma = 10,
                           y.name = "p", ar.method = "ols")
plot(res_bic_10)
```

The 10×10 BIC search in @fig-bic-10x10-subsets ranks candidate models across a broad parameter grid, with the top rows indicating the lowest BIC values. The most competitive specifications identify ARIMA(4,1,0) and ARIMA(6,1,0) as superior candidates, consistent with the visual diagnostics from the ACF and PACF plots.

#### 5×5 Search Space

```{r fig-bic-5x5-subsets}
#| label: fig-bic-5x5-subsets
#| fig-cap: "BIC subset selection — 5×5 search space. The restricted grid provides higher resolution for short-lag dependencies."
#| fig-height: 5

res_bic_5 <- armasubsets(y = diff_y_train, nar = 5, nma = 5,
                          y.name = "p", ar.method = "ols")
plot(res_bic_5)
```

Within the 5×5 restricted grid (@fig-bic-5x5-subsets), the BIC identifies ARIMA(0,1,4) as the top-performing model. The second-best introduces p-lag2 while retaining error-lag4, identifying ARIMA(2,1,4) as a secondary candidate.

### Candidate Model Selection

```{r stat-candidates}
#| label: stat-candidates

candidates_df <- data.frame(
  "Specification Source" = c("EACF", "BIC (5×5)", "BIC (10×10)", "ACF/PACF"),
  "Strategic Focus" = c(
    "Identifies stable vertices where the series is reduced to white noise.",
    "Targets lower-order mixed models using a restricted search space.",
    "Favours pure AR structures across a broader lag search.",
    "Directly addresses significant spikes observed at early lags in @fig-diff-acf-pacf."
  ),
  "Candidate Models" = c(
    "ARIMA(0,1,6), ARIMA(1,1,6), ARIMA(0,1,7)",
    "ARIMA(0,1,4), ARIMA(2,1,4)",
    "ARIMA(4,1,0), ARIMA(6,1,0)",
    "ARIMA(4,1,2), ARIMA(4,1,3), ARIMA(3,1,3)"
  ),
  check.names = FALSE
)

kbl(candidates_df,
    caption  = "Model Selection Framework — Candidate ARIMA Specifications for CBA Weekly Price Series",
    booktabs = TRUE,
    align    = "l") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE, width = "3cm") %>%
  column_spec(2, width = "9cm") %>%
  column_spec(3, width = "7cm")
```

These candidates span a spectrum from parsimonious low-order models to high-order specifications. This deliberate range tests whether added complexity resolves genuine structure or merely absorbs noise — a distinction that residual diagnostics and information criteria will determine.

## Model Fitting

All models are fitted to the raw (untransformed) original price series using `Arima()`. Both CSS and ML estimation are applied. CSS provides a computationally efficient approximation; ML maximises the exact likelihood. In a sample of this size (n ≈ `r length(y_train)`), the two methods should converge to nearly identical estimates — any material divergence is a signal of mis-specification or near-boundary parameters. `include.mean = FALSE` constrains the drift to zero, consistent with the no-drift assumption for CBA weekly price changes.

```{r fit-models}
#| label: fit-models

# ── EACF Vertex Models ────────────────────────────────────────────────────────
model_016_css <- Arima(y_train, order = c(0,1,6), method = "CSS")
model_016_ml  <- Arima(y_train, order = c(0,1,6), method = "ML")

model_116_css <- Arima(y_train, order = c(1,1,6), method = "CSS")
model_116_ml  <- Arima(y_train, order = c(1,1,6), method = "ML")

model_017_css <- Arima(y_train, order = c(0,1,7), method = "CSS")
model_017_ml  <- Arima(y_train, order = c(0,1,7), method = "ML")

# ── BIC (5×5) Models ──────────────────────────────────────────────────────────
model_014_css <- Arima(y_train, order = c(0,1,4), method = "CSS")
model_014_ml  <- Arima(y_train, order = c(0,1,4), method = "ML")

model_214_css <- Arima(y_train, order = c(2,1,4), method = "CSS")
model_214_ml  <- Arima(y_train, order = c(2,1,4), method = "ML")

# ── BIC (10×10) Models ────────────────────────────────────────────────────────
model_410_css <- Arima(y_train, order = c(4,1,0), method = "CSS")
model_410_ml  <- Arima(y_train, order = c(4,1,0), method = "ML")

model_610_css <- Arima(y_train, order = c(6,1,0), method = "CSS")
model_610_ml  <- Arima(y_train, order = c(6,1,0), method = "ML")

# ── ACF/PACF Candidates ───────────────────────────────────────────────────────
model_412_css <- Arima(y_train, order = c(4,1,2), method = "CSS")
model_412_ml  <- Arima(y_train, order = c(4,1,2), method = "ML")

model_413_css <- Arima(y_train, order = c(4,1,3), method = "CSS")
model_413_ml  <- Arima(y_train, order = c(4,1,3), method = "ML")

model_313_css <- Arima(y_train, order = c(3,1,3), method = "CSS")
model_313_ml  <- Arima(y_train, order = c(3,1,3), method = "ML")

cat("All candidate models fitted successfully under CSS and ML.\n")
```

### Parameter Estimation

The estimation of candidate models confirms that while complex structures capture specific signals, many alternatives fail due to statistical inefficiency and parameter redundancy. Tables are presented by identification source — CSS and ML columns appear consecutively for each model, making it straightforward to assess estimation convergence.

#### EACF Candidates: ARIMA(0,1,6), ARIMA(1,1,6), ARIMA(0,1,7)

```{r stat-coef-eacf}
#| label: stat-coef-eacf

make_coef_table(model_016_css, "ARIMA(0,1,6) — CSS Estimation")
make_coef_table(model_016_ml,  "ARIMA(0,1,6) — ML Estimation")
make_coef_table(model_116_css, "ARIMA(1,1,6) — CSS Estimation")
make_coef_table(model_116_ml,  "ARIMA(1,1,6) — ML Estimation")
make_coef_table(model_017_css, "ARIMA(0,1,7) — CSS Estimation")
make_coef_table(model_017_ml,  "ARIMA(0,1,7) — ML Estimation")
```

#### BIC 5×5 Grid Candidates: ARIMA(0,1,4), ARIMA(2,1,4)

```{r stat-coef-bic5}
#| label: stat-coef-bic5

make_coef_table(model_014_css, "ARIMA(0,1,4) — CSS Estimation")
make_coef_table(model_014_ml,  "ARIMA(0,1,4) — ML Estimation")
make_coef_table(model_214_css, "ARIMA(2,1,4) — CSS Estimation")
make_coef_table(model_214_ml,  "ARIMA(2,1,4) — ML Estimation")
```

#### BIC 10×10 Grid Candidates: ARIMA(4,1,0), ARIMA(6,1,0)

```{r stat-coef-bic10}
#| label: stat-coef-bic10

make_coef_table(model_410_css, "ARIMA(4,1,0) — CSS Estimation")
make_coef_table(model_410_ml,  "ARIMA(4,1,0) — ML Estimation")
make_coef_table(model_610_css, "ARIMA(6,1,0) — CSS Estimation")
make_coef_table(model_610_ml,  "ARIMA(6,1,0) — ML Estimation")
```

#### ACF/PACF Diagnostic Candidates: ARIMA(4,1,2), ARIMA(4,1,3), ARIMA(3,1,3)

```{r stat-coef-acfpacf}
#| label: stat-coef-acfpacf

make_coef_table(model_412_css, "ARIMA(4,1,2) — CSS Estimation")
make_coef_table(model_412_ml,  "ARIMA(4,1,2) — ML Estimation")
make_coef_table(model_413_css, "ARIMA(4,1,3) — CSS Estimation")
make_coef_table(model_413_ml,  "ARIMA(4,1,3) — ML Estimation")
make_coef_table(model_313_css, "ARIMA(3,1,3) — CSS Estimation")
make_coef_table(model_313_ml,  "ARIMA(3,1,3) — ML Estimation")
```

The CSS and ML estimation results for ARIMA(3,1,3) and ARIMA(4,1,3) highlight a "sparse" dependency structure where higher-order lags are more informative than intermediate ones.

**Strong structural signals.** Both models capture highly significant dependencies at the third and fourth lags. In the ARIMA(3,1,3) specification, `ar3` and `ma3` are both significant at the p \< 0.001 level.

**Validation of lag 4.** The ARIMA(4,1,3) model successfully validates the "lag 4" signal identified in previous diagnostics, with the `ar4` coefficient showing statistical significance.

**Internal inefficiency.** Despite capturing these signals, both models suffer from non-significant intermediate parameters. Specifically, the second-order lags (`ar2` and `ma2`) consistently fail to reach significance across both CSS and ML estimations, with p-values often exceeding 0.33. This sparse structure is a known feature of financial return series where memory operates at specific lag intervals rather than uniformly across all orders.

### Information Criteria

```{r stat-aic-bic-ranking}
#| label: stat-aic-bic-ranking

all_models_ml <- list(
  "ARIMA(0,1,6)" = model_016_ml,
  "ARIMA(1,1,6)" = model_116_ml,
  "ARIMA(0,1,7)" = model_017_ml,
  "ARIMA(0,1,4)" = model_014_ml,
  "ARIMA(2,1,4)" = model_214_ml,
  "ARIMA(4,1,0)" = model_410_ml,
  "ARIMA(6,1,0)" = model_610_ml,
  "ARIMA(4,1,2)" = model_412_ml,
  "ARIMA(4,1,3)" = model_413_ml,
  "ARIMA(3,1,3)" = model_313_ml
)

ic_df <- data.frame(
  Model = names(all_models_ml),
  AIC   = round(sapply(all_models_ml, AIC), 3),
  BIC   = round(sapply(all_models_ml, BIC), 3)
)
ic_df_sorted        <- ic_df[order(ic_df$BIC), ]
rownames(ic_df_sorted) <- NULL

knitr::kable(ic_df_sorted,
  caption = "Information criteria — all candidate models sorted by BIC (ML estimation).",
  align   = "c")
```

The table presents a systematic ranking of candidate models, comparing their ability to balance goodness-of-fit against model parsimony.

**BIC** imposes a more stringent penalty for model complexity (logarithm of the sample size) than AIC, prioritising the minimisation of over-parameterisation risk. **ARIMA(3,1,3)** achieves the lowest BIC (1360.501), making it the most efficient specification within the candidate grid. Despite the statistical insignificance of intermediate lags (`ar2`/`ma2`) noted during parameter estimation, the p = 3, q = 3 structure is identified as the most efficient balance for the CBA data. ARIMA(0,1,4) follows closely (BIC: 1361.071), suggesting that a simplified moving-average structure captures the fourth-lag dependency nearly as efficiently.

**AIC** is generally more lenient toward complexity, as it seeks to minimise information loss rather than identify the "true" model dimension. **ARIMA(4,1,3)** achieves the lowest AIC (1334.435), favouring the inclusion of the 4th autoregressive lag (`ar4`), which was previously identified as a significant signal in the BIC 10×10 search.

While ARIMA(4,1,3) offers the best overall fit by AIC, ARIMA(3,1,3) is the superior choice for a robust analytical report due to its balance of complexity and performance. It achieves the lowest BIC, which is the most critical metric for avoiding over-parameterisation in financial time series, and achieves nearly the same explanatory power without the additional penalty of a fourth autoregressive parameter.

### Accuracy Metrics

```{r stat-accuracy-metrics}
#| label: stat-accuracy-metrics

# Compute in-sample accuracy for all fitted ML models
accuracy_014_ml <- accuracy(model_014_ml)[1:7]
accuracy_214_ml <- accuracy(model_214_ml)[1:7]
accuracy_410_ml <- accuracy(model_410_ml)[1:7]
accuracy_610_ml <- accuracy(model_610_ml)[1:7]
accuracy_412_ml <- accuracy(model_412_ml)[1:7]
accuracy_413_ml <- accuracy(model_413_ml)[1:7]
accuracy_313_ml <- accuracy(model_313_ml)[1:7]
accuracy_016_ml <- accuracy(model_016_ml)[1:7]
accuracy_116_ml <- accuracy(model_116_ml)[1:7]
accuracy_017_ml <- accuracy(model_017_ml)[1:7]

df_accuracy <- data.frame(
  rbind(
    accuracy_014_ml, accuracy_214_ml,
    accuracy_410_ml, accuracy_610_ml,
    accuracy_412_ml, accuracy_413_ml, accuracy_313_ml,
    accuracy_016_ml, accuracy_116_ml, accuracy_017_ml
  )
)
colnames(df_accuracy) <- c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE", "ACF1")
rownames(df_accuracy) <- c(
  "ARIMA(0,1,4)-ML", "ARIMA(2,1,4)-ML",
  "ARIMA(4,1,0)-ML", "ARIMA(6,1,0)-ML",
  "ARIMA(4,1,2)-ML", "ARIMA(4,1,3)-ML", "ARIMA(3,1,3)-ML",
  "ARIMA(0,1,6)-ML", "ARIMA(1,1,6)-ML", "ARIMA(0,1,7)-ML"
)

# Composite rank (lower is better on all four metrics)
df_accuracy$RMSE_Rank <- rank(df_accuracy$RMSE)
df_accuracy$MAE_Rank  <- rank(df_accuracy$MAE)
df_accuracy$MAPE_Rank <- rank(df_accuracy$MAPE)
df_accuracy$MASE_Rank <- rank(df_accuracy$MASE)
df_accuracy$Total_Rank <- rowSums(
  df_accuracy[, c("RMSE_Rank", "MAE_Rank", "MAPE_Rank", "MASE_Rank")]
)
df_accuracy_sorted <- df_accuracy[order(df_accuracy$Total_Rank), ]

knitr::kable(
  round(df_accuracy_sorted[, c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE", "ACF1")], 4),
  caption = "In-sample accuracy metrics for all candidate models (ML estimation), sorted by composite rank.",
  align   = "c"
)
```

The analysis of in-sample accuracy metrics reinforces the selection of ARIMA(3,1,3) as the most robust candidate. It achieves the lowest Mean Error (ME = 0.3253) and Mean Absolute Error (MAE = 2.3748), demonstrating superior accuracy in predicting the magnitude of weekly price movements. It also maintains the lowest Mean Absolute Scaled Error (MASE = 0.1132), confirming its efficiency relative to a naive baseline.

## Overparameterisation Check

To confirm that ARIMA(3,1,3) is not under-specified, we test one-order augmentations: ARIMA(4,1,4) and ARIMA(3,1,4).

### Parameter Estimation: Over-parameterisation Check

```{r stat-overpara}
#| label: stat-overpara

model_414_css <- Arima(y_train, order = c(4,1,4), method = "CSS")
model_414_ml  <- Arima(y_train, order = c(4,1,4), method = "ML")
model_314_css <- Arima(y_train, order = c(3,1,4), method = "CSS")
model_314_ml  <- Arima(y_train, order = c(3,1,4), method = "ML")

# Reference models for comparison
make_coef_table(model_413_css, "ARIMA(4,1,3) — CSS Estimation")
make_coef_table(model_413_ml,  "ARIMA(4,1,3) — ML Estimation")
make_coef_table(model_313_css, "ARIMA(3,1,3) — CSS Estimation")
make_coef_table(model_313_ml,  "ARIMA(3,1,3) — ML Estimation")

# Augmented models
make_coef_table(model_414_css, "ARIMA(4,1,4) — CSS Estimation")
make_coef_table(model_414_ml,  "ARIMA(4,1,4) — ML Estimation")
make_coef_table(model_314_css, "ARIMA(3,1,4) — CSS Estimation")
make_coef_table(model_314_ml,  "ARIMA(3,1,4) — ML Estimation")
```

### Information Criteria: Over-parameterisation Check

```{r stat-aic-bic-ranking-overpara}
#| label: stat-aic-bic-ranking-overpara

all_models_ml_overpara <- list(
  "ARIMA(0,1,6)" = model_016_ml,
  "ARIMA(1,1,6)" = model_116_ml,
  "ARIMA(0,1,7)" = model_017_ml,
  "ARIMA(0,1,4)" = model_014_ml,
  "ARIMA(2,1,4)" = model_214_ml,
  "ARIMA(4,1,0)" = model_410_ml,
  "ARIMA(6,1,0)" = model_610_ml,
  "ARIMA(4,1,2)" = model_412_ml,
  "ARIMA(4,1,3)" = model_413_ml,
  "ARIMA(3,1,3)" = model_313_ml,
  "ARIMA(4,1,4) [augmented]" = model_414_ml,
  "ARIMA(3,1,4) [augmented]" = model_314_ml
)

ic_df_overpara <- data.frame(
  Model = names(all_models_ml_overpara),
  AIC   = round(sapply(all_models_ml_overpara, AIC), 3),
  BIC   = round(sapply(all_models_ml_overpara, BIC), 3)
)
ic_df_overpara_sorted        <- ic_df_overpara[order(ic_df_overpara$BIC), ]
rownames(ic_df_overpara_sorted) <- NULL

knitr::kable(ic_df_overpara_sorted,
  caption = "Information criteria including augmented models — sorted by BIC (ML estimation).",
  align   = "c")
```

Adding a fourth moving-average parameter (ARIMA(3,1,4)) or a combined fourth AR and MA parameter (ARIMA(4,1,4)) does not improve BIC relative to ARIMA(3,1,3). This confirms that the additional parameters absorb noise rather than genuine signal. ARIMA(3,1,3) remains the BIC-optimal specification even after augmentation.

### Accuracy Metrics: Over-parameterisation Check

```{r stat-accuracy-metrics-overpara}
#| label: stat-accuracy-metrics-overpara

accuracy_414_ml <- accuracy(model_414_ml)[1:7]
accuracy_314_ml <- accuracy(model_314_ml)[1:7]

df_accuracy_overpara <- data.frame(
  rbind(
    accuracy_014_ml, accuracy_214_ml,
    accuracy_410_ml, accuracy_610_ml,
    accuracy_412_ml, accuracy_413_ml, accuracy_313_ml,
    accuracy_016_ml, accuracy_116_ml, accuracy_017_ml,
    accuracy_414_ml, accuracy_314_ml
  )
)
colnames(df_accuracy_overpara) <- c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE", "ACF1")
rownames(df_accuracy_overpara) <- c(
  "ARIMA(0,1,4)-ML", "ARIMA(2,1,4)-ML",
  "ARIMA(4,1,0)-ML", "ARIMA(6,1,0)-ML",
  "ARIMA(4,1,2)-ML", "ARIMA(4,1,3)-ML", "ARIMA(3,1,3)-ML",
  "ARIMA(0,1,6)-ML", "ARIMA(1,1,6)-ML", "ARIMA(0,1,7)-ML",
  "ARIMA(4,1,4)-ML [augmented]", "ARIMA(3,1,4)-ML [augmented]"
)

df_accuracy_overpara$RMSE_Rank <- rank(df_accuracy_overpara$RMSE)
df_accuracy_overpara$MAE_Rank  <- rank(df_accuracy_overpara$MAE)
df_accuracy_overpara$MAPE_Rank <- rank(df_accuracy_overpara$MAPE)
df_accuracy_overpara$MASE_Rank <- rank(df_accuracy_overpara$MASE)
df_accuracy_overpara$Total_Rank <- rowSums(
  df_accuracy_overpara[, c("RMSE_Rank", "MAE_Rank", "MAPE_Rank", "MASE_Rank")]
)
df_accuracy_overpara_sorted <- df_accuracy_overpara[order(df_accuracy_overpara$Total_Rank), ]

knitr::kable(
  round(df_accuracy_overpara_sorted[, c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE", "ACF1")], 4),
  caption = "In-sample accuracy metrics including augmented models (ML estimation), sorted by composite rank.",
  align   = "c"
)
```

ARIMA(3,1,3) maintains its ranking even when augmented models are included in the accuracy comparison. The augmented specifications offer no improvement in MASE or MAE, and their higher BIC values confirm that the additional parameters do not capture genuine data structure. The over-parameterisation check formally validates the ARIMA(3,1,3) selection.

## Residual Diagnostics

The ARIMA(3,1,3) model is successfully specified in terms of its mean structure. The Ljung-Box p-values (all above 0.05 across lags 1–40) and the ACF of residuals (no significant spikes) confirm that the residuals constitute independent white noise — the residual autocorrelation structure has been fully absorbed by the model.

```{r fig-residual-313}
#| label: fig-residual-313-overpara
#| fig-cap: "ARIMA(3,1,3) ML — residual diagnostics reproduced for direct comparison against augmented models. The Ljung-Box panel confirms white noise residuals are maintained."
#| fig-height: 12

plot_residual_diagnostics(model_313_ml, "ARIMA(3,1,3) ML", lag.max = 40)
```

However, the normality assumption is violated. The QQ plot reveals heavy tails, and the histogram shows pronounced kurtosis relative to the Gaussian reference. These outliers are concentrated in the 2025–2026 period and likely represent idiosyncratic market shocks — abrupt repricing events driven by unexpected RBA decisions or global credit stress — that a linear ARIMA model cannot fully capture. This violation does not invalidate the model's mean-structure specification but implies that prediction intervals should be interpreted with caution, as they assume Gaussian residuals.

## Model Selection

### ARIMA(3,1,3) — ML Fitted Values

```{r fig-fitted-313}
#| label: fig-fitted-313
#| fig-cap: "ARIMA(3,1,3) ML fitted values (blue) overlaid on the observed CBA training price series (black). Close tracking across the full training window indicates effective in-sample fit."
#| fig-height: 5

plot_ml_fitted(model_313_ml, y_train, spec = "ARIMA(3,1,3)")
```

The fitted-value overlay in @fig-fitted-313 demonstrates that ARIMA(3,1,3) ML closely tracks the observed training series throughout the full sample period. The fitted values capture both the sustained upward trend (reflecting the integrated nature of the model at d = 1) and the localised week-to-week fluctuations attributable to the AR(3) and MA(3) components. No systematic divergence or structural drift is evident, which is consistent with a well-specified model.

### Final Model Selection and Justification

```{r stat-best-model}
#| label: stat-best-model

best_model      <- model_313_ml
best_model_name <- "ARIMA(3,1,3) — ML"

cat("Selected model:", best_model_name, "\n\n")
print(summary(best_model))
```

The convergence of evidence across all specification, estimation, and diagnostic stages leads to the selection of **ARIMA(3,1,3) estimated by ML** as the final model.

**BIC optimality.** ARIMA(3,1,3) achieves the lowest BIC (1360.501) across the full candidate grid, including the augmented over-parameterisation models. BIC is the decisive criterion here because it penalises complexity proportional to log(n), guarding against the noise absorption that inflates in-sample fit without improving out-of-sample accuracy.

**In-sample accuracy.** ARIMA(3,1,3) achieves the lowest ME (0.3253) and MAE (2.3748) among all candidates, and the lowest MASE (0.1132) relative to the naive benchmark. These metrics confirm that the model translates its structural specification into measurable forecasting efficiency.

**Residual white noise.** The Ljung-Box test (all p-values \> 0.05 across lags 1–40) and the flat ACF of residuals confirm that the model has fully absorbed the autocorrelative structure of the training series. No residual information remains exploitable.

**Parsimony over ARIMA(4,1,3).** Although ARIMA(4,1,3) achieves a lower AIC (1334.435), its additional `ar4` parameter incurs a meaningful BIC penalty without delivering a commensurate improvement in residual diagnostics or accuracy metrics. The fourth autoregressive lag captures real signal identified in the BIC 10×10 grid, but the marginal gain does not justify the added complexity under the BIC criterion.

**Sparse structure acknowledged.** The insignificance of intermediate lags (`ar2`/`ma2`) is a known feature of financial return series where memory operates at specific lag intervals. This "sparse" dependency structure does not invalidate the model — it reflects the genuine autocorrelative geometry of CBA weekly returns rather than mis-specification.

**Residual non-normality noted.** The Shapiro-Wilk test rejects normality for the standardised residuals, attributable to extreme observations in the 2025–2026 period. This is a limitation of the Gaussian ARIMA framework and should be addressed by a GARCH extension if precise prediction interval coverage is required. It does not affect the validity of the point forecasts or the white-noise criterion.

## Forecasting

We evaluate the selected ARIMA(3,1,3) model alongside a comprehensive suite of alternative forecasting methods — benchmarks, exponential smoothing, automatic ARIMA, and TBATS — to provide a complete picture of forecast performance on the 12-week test set. All methods are applied to the original (untransformed) price series.

```{r fc-setup}
#| label: fc-setup

h <- length(y_test)
cat("Forecast horizon h =", h, "weeks\n")
```

### ARIMA(3,1,3) Forecast

```{r forecast-table}
#| label: forecast-table

fc_arima <- forecast(best_model, h = h)

fc_df <- data.frame(
  Week     = seq_along(fc_arima$mean),
  Forecast = round(as.numeric(fc_arima$mean),           2),
  Lo_80    = round(as.numeric(fc_arima$lower[, "80%"]), 2),
  Hi_80    = round(as.numeric(fc_arima$upper[, "80%"]), 2),
  Lo_95    = round(as.numeric(fc_arima$lower[, "95%"]), 2),
  Hi_95    = round(as.numeric(fc_arima$upper[, "95%"]), 2),
  Actual   = round(as.numeric(y_test),                  2)
)

knitr::kable(fc_df,
  caption   = paste0("12-week ahead forecast vs actual CBA.AX prices — ", best_model_name, "."),
  align     = "c",
  col.names = c("Week", "Forecast (AUD)", "Lo 80%", "Hi 80%",
                "Lo 95%", "Hi 95%", "Actual (AUD)"))
```

```{r fig-fc-arima}
#| label: fig-fc-arima
#| fig-cap: "ARIMA(3,1,3) ML — 12-week forecast. Shaded regions show 80% and 95% prediction intervals; red line shows actual test-set prices."
#| fig-height: 5

plot_forecast_simple(fc_arima, y_train, y_test, best_model_name)
```

**Convergence to flat trajectory** (@fig-fc-arima). The ARIMA(3,1,3) model provides short-horizon corrections attributable to the AR(3) and MA(3) components; beyond three to four periods, the expected future price change approaches zero and the forecast converges to a flat level. This is an honest reflection of the near-random-walk nature of CBA prices rather than a model failure — it correctly communicates that the historical autocorrelative structure does not support directional price prediction at extended horizons.

**Widening intervals.** Prediction intervals widen symmetrically as the forecast horizon extends, reflecting the linear accumulation of uncertainty in an I(1) process. Given the residual non-normality identified above, the nominal 95% coverage may understate the true uncertainty around extreme shock events.

### Benchmark Forecasts

```{r fc-benchmarks}
#| label: fc-benchmarks

fc_bench <- benchmark_forecasts(y_train, y_test)
```

```{r fig-fc-naive}
#| label: fig-fc-naive
#| fig-cap: "Naive forecast: the last observed training price carried forward as a flat line. This represents the simplest possible forecast and serves as the baseline every model must outperform."
#| fig-height: 4

plot_forecast_simple(fc_bench$Naive, y_train, y_test, "Naive")
```

```{r fig-fc-drift}
#| label: fig-fc-drift
#| fig-cap: "Drift forecast: linear extrapolation of the historical average weekly change. The drift captures the long-run upward trend in CBA prices but ignores any mean-reversion or shock dynamics."
#| fig-height: 4

plot_forecast_simple(fc_bench$Drift, y_train, y_test, "Drift")
```

```{r fig-fc-mean}
#| label: fig-fc-mean
#| fig-cap: "Mean forecast: the historical training average projected as a constant. This assumes CBA prices will immediately revert to the long-run mean — an assumption violated by the unit-root evidence."
#| fig-height: 4

plot_forecast_simple(fc_bench$Mean, y_train, y_test, "Mean")
```

The three benchmark methods represent simple but important baselines. The Naive method (last observation carried forward) is theoretically equivalent to a random walk forecast and should perform competitively with ARIMA models on near-random-walk data. The Drift method extrapolates the historical average weekly change, capturing the training-period upward trend. The Mean method projects the historical average and will perform poorly whenever the test-period price level deviates materially from the training mean.

### Exponential Smoothing (ETS)

```{r fc-ets}
#| label: fc-ets

fc_ets_result <- ets_forecast(y_train, h = h)
cat("ETS model selected:", fc_ets_result$model$method, "\n")
```

```{r fig-fc-ets}
#| label: fig-fc-ets
#| fig-cap: "ETS forecast with automatically selected error, trend, and seasonal components. ETS adapts to the local level and trend in the training data through exponentially declining weights on past observations."
#| fig-height: 5

plot_forecast_simple(fc_ets_result$forecast, y_train, y_test,
                     model_name = paste0("ETS(", fc_ets_result$model$method, ")"))
```

Exponential Smoothing State Space (ETS) models assign exponentially declining weights to past observations, with the rate of decay governed by smoothing parameters. For a trending financial series like CBA, ETS typically selects an additive trend component (model type A,A,N), allowing it to project the local upward trend forward. The `ets()` function's automatic selection process chooses the component structure that minimises the training-set information criterion, providing a data-driven alternative to the manual ARIMA specification.

### Automatic ARIMA

```{r fc-autoarima}
#| label: fc-autoarima

fc_auto_result <- auto_arima_forecast(y_train, h = h)
cat("auto.arima selected:", as.character(fc_auto_result$model), "\n")
```

```{r fig-fc-autoarima}
#| label: fig-fc-autoarima
#| fig-cap: "auto.arima() forecast using an exhaustive (non-stepwise, exact likelihood) search over ARIMA order combinations. This serves as an algorithmic benchmark against the manually specified ARIMA(3,1,3)."
#| fig-height: 5

plot_forecast_simple(fc_auto_result$forecast, y_train, y_test,
                     model_name = paste0("auto.arima: ",
                                         as.character(fc_auto_result$model)))
```

`auto.arima()` with `stepwise = FALSE` and `approximation = FALSE` performs an exhaustive search over ARIMA order combinations using exact likelihood. Comparing its selected order against the manually specified ARIMA(3,1,3) validates the model specification process: if `auto.arima()` selects the same model, the manual analysis is confirmed; if it selects a different specification, the discrepancy warrants investigation.

### TBATS

```{r fc-tbats}
#| label: fc-tbats

fc_tbats_result <- tbats_forecast(y_train, h = h)
cat("TBATS model fitted successfully.\n")
```

```{r fig-fc-tbats}
#| label: fig-fc-tbats
#| fig-cap: "TBATS forecast — handles complex seasonal patterns including multiple seasonal periods. For weekly financial data with frequency = 52, TBATS may capture annual trading cycles that ARIMA cannot model."
#| fig-height: 5

plot_forecast_simple(fc_tbats_result$forecast, y_train, y_test, "TBATS")
```

TBATS (Trigonometric, Box-Cox, ARMA errors, Trend, Seasonal) decomposes the series into trend, seasonal, and error components, handling multiple seasonal periods and non-integer frequencies. For CBA weekly data with frequency = 52, TBATS can model any annual trading cycle (e.g., year-end institutional rebalancing) that ARIMA treats as noise. If TBATS achieves materially better test-set accuracy than ARIMA, this suggests a seasonal component with practical significance for CBA price dynamics.

### Forecast Comparison

```{r fig-fc-comparison}
#| label: fig-fc-comparison
#| fig-cap: "All forecasting methods overlaid for direct visual comparison. The red dashed line shows actual test-set prices. Methods that track the actual series most closely in direction and magnitude achieve the best out-of-sample accuracy."
#| fig-height: 6

all_forecasts <- list(
  "ARIMA(3,1,3)" = fc_arima,
  "Naive"        = fc_bench$Naive,
  "Drift"        = fc_bench$Drift,
  "Mean"         = fc_bench$Mean,
  "ETS"          = fc_ets_result$forecast,
  "auto.arima"   = fc_auto_result$forecast,
  "TBATS"        = fc_tbats_result$forecast
)

plot_forecast_comparison(all_forecasts, y_train, y_test,
  title = "CBA.AX — All Forecast Methods vs Actual (12-Week Test Set)")
```

@fig-fc-comparison provides the definitive visual comparison of all methods against the actual test-set prices. Methods that diverge substantially from the actual series — particularly any that project the training trend steeply upward while actual prices are flat or declining — are demonstrating over-reliance on historical momentum, which makes directional forecasting of near-random-walk equity prices so difficult. Methods that remain close to the last observed price (Naive, ARIMA) are implicitly acknowledging the random-walk structure and will typically outperform more opinionated methods on error metrics.

### Accuracy Comparison

```{r stat-forecast-accuracy}
#| label: stat-forecast-accuracy

compare_forecasts(all_forecasts, y_test,
  caption = "Out-of-sample forecast accuracy — all methods ranked by RMSE (12-week test set).")
```

```{r stat-forecast-table-arima}
#| label: stat-forecast-table-arima

fc_df_final <- data.frame(
  Week     = seq_along(fc_arima$mean),
  Forecast = round(as.numeric(fc_arima$mean),           2),
  Lo_80    = round(as.numeric(fc_arima$lower[, "80%"]), 2),
  Hi_80    = round(as.numeric(fc_arima$upper[, "80%"]), 2),
  Lo_95    = round(as.numeric(fc_arima$lower[, "95%"]), 2),
  Hi_95    = round(as.numeric(fc_arima$upper[, "95%"]), 2),
  Actual   = round(as.numeric(y_test),                  2)
)

knitr::kable(fc_df_final,
  caption   = paste0("12-week forecast vs actual CBA.AX prices — ", best_model_name, "."),
  align     = "c",
  col.names = c("Week", "Forecast (AUD)", "Lo 80%", "Hi 80%",
                "Lo 95%", "Hi 95%", "Actual (AUD)"))
```

The accuracy table ranks all seven methods by out-of-sample RMSE against the 12-week test set. For a near-random-walk series, the Naive forecast is the natural benchmark — any method that fails to outperform the Naive on RMSE has not extracted useful predictive signal from the historical data beyond the last observation. ARIMA(3,1,3) is expected to perform similarly to the Naive at longer horizons (both converge to a flat projection), but its short-horizon corrections from the AR(3) and MA(3) terms may yield a marginal edge in the first three to four weeks of the test set.

## Conclusion

This analysis applied the full Box–Jenkins ARIMA workflow and a comprehensive suite of alternative forecasting methods to the weekly adjusted closing price of Commonwealth Bank of Australia (CBA.AX) over a five-year sample.

**Transformation (rejected).** The Box-Cox analysis (@fig-boxcox-profile) supported a log transformation statistically, but the applied transformation produced no material improvement in normality or variance stability relative to the original series (@fig-bc-norm, @fig-bc-compare). In accordance with parsimony, the original price series was retained for modelling.

**Integration order (d = 1).** The formal stationarity battery — ADF, PP, and KPSS — unanimously confirmed a unit root in the raw series (ADF p = `r round(stat_raw$result$p.value, 4)`) and stationarity after one round of differencing. CBA's price path incorporates each new shock — RBA decisions, global credit events, domestic housing data — permanently into the price level.

**Model specification.** The ACF/PACF (@fig-diff-acf-pacf), EACF, and dual BIC analysis (@fig-bic-10x10-subsets, @fig-bic-5x5-subsets) generated a candidate set spanning parsimonious low-order models through complex higher-order specifications. The diagnostic evidence consistently pointed toward a (3,1,3) or (4,1,3) structure, with the third and fourth lags carrying the primary autocorrelative signal.

**Model selection.** ARIMA(3,1,3) ML was selected as the final model on the convergence of three criteria: the lowest BIC (1360.501) across all candidate and augmented models, the best composite in-sample accuracy (ME = 0.3253, MAE = 2.3748, MASE = 0.1132), and the confirmation of white-noise residuals via Ljung-Box testing. The sparse structure — with insignificant intermediate lags (`ar2`/`ma2`) — reflects the genuine autocorrelative geometry of CBA weekly returns rather than model mis-specification. Although ARIMA(4,1,3) achieved a lower AIC, its marginal additional complexity was not rewarded by BIC or accuracy metrics, and the over-parameterisation check formally confirmed that the added parameter absorbs noise.

**Residual adequacy.** The ARIMA(3,1,3) specification passes the Ljung-Box white-noise criterion across all tested lags. Residual non-normality, attributable to extreme observations in 2025–2026, reflects the known fat-tail property of financial returns — a limitation of the Gaussian ARIMA framework addressed by GARCH extensions rather than higher ARIMA orders.

**Forecasting.** The multi-method comparison (@fig-fc-comparison) demonstrates the fundamental difficulty of forecasting near-random-walk stock prices. Any method that does not substantially outperform the Naive benchmark on out-of-sample RMSE has not extracted exploitable information from the historical price series — an empirical confirmation of weak-form market efficiency for CBA.AX.

**Limitations and extensions.** CBA weekly returns exhibit volatility clustering visible in @fig-diff-series that no ARIMA model can capture. A GARCH(1,1) or EGARCH extension would model conditional heteroscedasticity and produce more accurate prediction intervals. Incorporating exogenous variables (ARIMAX) — the RBA cash rate, ASX 200 index, CBA's net interest margin — would leverage macroeconomic theory to improve point forecast accuracy beyond what univariate price history provides.

## References

Cryer, J.D. & Chan, K.-S. (2008). *Time Series Analysis with R*. Springer.

Box, G.E.P. & Jenkins, G.M. (1970). *Time Series Analysis: Forecasting and Control*. Holden-Day.

Hyndman, R.J. & Athanasopoulos, G. (2021). *Forecasting: Principles and Practice* (3rd ed.). OTexts.

Demirhan, H. (2026). MATH1318/2204 Module 7 — Model Diagnostics and Forecasting. RMIT University.
