---
title: "Forecasting ASX Stocks"
subtitle: "Time Series Analysis of CBA Daily Adjusted Closing Price"
author: "Yemunn Soe"
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: true
    fig-cap-location: bottom
    theme: flatly
---

```{r setup}
#| label: setup
#| include: false
#| cache: false

library(tidyquant)
library(dplyr)
library(TSA)
library(forecast)
library(tseries)
library(lmtest)
library(ggplot2)
library(purrr)
library(magrittr)
library(knitr)
library(kableExtra)
```

```{r functions}
#| label: functions
#| include: false

# ══════════════════════════════════════════════════════════════════════════════
# HELPER FUNCTIONS
# A. Plotting helpers     (1–3, 8, 20–21)
# B. Formal test helpers  (4–5)
# C. Residual diagnostics (9–11)
# D. Output helpers       (7)
# E. Forecasting          (12–19)
# ══════════════════════════════════════════════════════════════════════════════

# ── Function 1: Time series plot with optional mean line ──────────────────────
plot_ts <- function(ts_data, main, ylab = "Price", add_mean = FALSE) {
  plot(ts_data, type = "l", ylab = ylab, xlab = "Time",
       cex.main = 0.9, main = main)
  if (add_mean) {
    abline(h = mean(ts_data, na.rm = TRUE),
           col = "blue", lty = 2, lwd = 1.5)
  }
}

# ── Function 2: Side-by-side ACF and PACF plots ───────────────────────────────
plot_acf_pacf <- function(ts_data, acf_title, pacf_title, lag.max = 40) {
  par(mfrow = c(1, 2), cex.main = 0.9)
  acf(ts_data,  lag.max = lag.max, main = acf_title,  ylab = "ACF",  xlab = "Lag")
  pacf(ts_data, lag.max = lag.max, main = pacf_title, ylab = "PACF", xlab = "Lag")
  par(mfrow = c(1, 1))
}

# ── Function 3: Side-by-side QQ plot and Histogram ────────────────────────────
plot_norm <- function(ts_data, qq_title, hist_title, xlab = "Value") {
  par(mfrow = c(1, 2), cex.main = 0.9)
  qqnorm(ts_data, main = qq_title, pch = 16, cex = 0.7)
  qqline(ts_data, col = "blue", lty = 2, lwd = 1.5)
  hist(ts_data, xlab = xlab, main = hist_title,
       border = "white", breaks = 20)
  curve(
    dnorm(x,
          mean = mean(ts_data, na.rm = TRUE),
          sd   = sd(ts_data,   na.rm = TRUE)) *
      length(ts_data) *
      diff(hist(ts_data, plot = FALSE, breaks = 20)$breaks)[1],
    add = TRUE, col = "blue", lwd = 2
  )
  par(mfrow = c(1, 1))
}

# ── Function 4: Shapiro-Wilk normality test with kable output ─────────────────
check_normality <- function(ts_data, caption_text) {
  sw <- shapiro.test(as.numeric(ts_data))
  p_formatted <- format(sw$p.value, scientific = TRUE, digits = 2)
  df <- data.frame(
    "Test Statistic (W)" = round(sw$statistic, 4),
    "p-value"            = p_formatted,
    check.names          = FALSE
  )
  list(
    result = sw,
    table  = knitr::kable(df, caption = caption_text,
                          align = "c", row.names = FALSE)
  )
}

# ── Function 5: Full stationarity test battery (ADF + PP + KPSS) ──────────────
check_stationarity <- function(ts_data, table_title) {
  adf_res  <- adf.test(ts_data,  alternative = "stationary")
  pp_res   <- pp.test(ts_data,   alternative = "stationary")
  kpss_res <- kpss.test(ts_data, null        = "Trend")

  adf_v  <- ifelse(adf_res$p.value  < 0.05,
                   "Reject H0 → Stationary",
                   "Fail to Reject H0 → Non-stationary")
  pp_v   <- ifelse(pp_res$p.value   < 0.05,
                   "Reject H0 → Stationary",
                   "Fail to Reject H0 → Non-stationary")
  kpss_v <- ifelse(kpss_res$p.value > 0.05,
                   "Fail to Reject H0 → Stationary",
                   "Reject H0 → Non-stationary")

  df <- data.frame(
    "Test"      = c("Augmented Dickey-Fuller (ADF)",
                    "Phillips-Perron (PP)", "KPSS"),
    "H0"        = c("Unit root (non-stationary)",
                    "Unit root (non-stationary)",
                    "Series is stationary"),
    "Statistic" = c(round(unname(adf_res$statistic),  4),
                    round(unname(pp_res$statistic),   4),
                    round(unname(kpss_res$statistic), 4)),
    "p-value"   = c(format(adf_res$p.value,  scientific = TRUE, digits = 2),
                    format(pp_res$p.value,   scientific = TRUE, digits = 2),
                    format(kpss_res$p.value, scientific = TRUE, digits = 2)),
    "Decision"  = c(adf_v, pp_v, kpss_v),
    check.names = FALSE
  )
  list(
    adf   = adf_res, pp = pp_res, kpss = kpss_res,
    table = knitr::kable(df, caption = table_title, align = "llccc")
  )
}

# ── Function 7: Coefficient significance table ────────────────────────────────
make_coef_table <- function(model, caption_text) {
  ct <- lmtest::coeftest(model)
  df <- data.frame(
    Coefficient = rownames(ct),
    Estimate    = round(ct[, 1], 5),
    Std.Error   = round(ct[, 2], 5),
    z.value     = round(ct[, 3], 4),
    p.value     = round(ct[, 4], 4),
    Sig         = ifelse(ct[, 4] < 0.001, "***",
                  ifelse(ct[, 4] < 0.01,  "**",
                  ifelse(ct[, 4] < 0.05,  "*",
                  ifelse(ct[, 4] < 0.1,   ".", ""))))
  )
  knitr::kable(df, caption = caption_text, align = "c", row.names = FALSE)
}

# ── Function 8: CSS vs ML fitted value overlay plots ──────────────────────────
plot_css_ml <- function(css_model, ml_model, ts_data, spec) {
  par(mfrow = c(1, 2), cex.main = 0.85, mar = c(4, 4, 3, 2))
  obs_col <- adjustcolor("black", alpha.f = 0.7)
  for (i in 1:2) {
    mod   <- if (i == 1) css_model else ml_model
    label <- if (i == 1) paste(spec, "(CSS)") else paste(spec, "(ML)")
    plot(ts_data, type = "l", ylab = "Price (AUD)", xlab = "",
         main = label, cex.main = 0.85, col = obs_col)
    lines(fitted(mod), col = "blue", lwd = 2)
    legend("topleft", legend = c("Observed", "Fitted"),
           col = c(obs_col, "blue"), lty = 1, lwd = c(1, 2),
           cex = 0.65, bty = "n")
  }
  par(mfrow = c(1, 1))
}

# ── Function 9: Residual visual panel (2×2 grid) ──────────────────────────────
plot_residuals_panel <- function(model, model_name, lag.max = 40) {
  res <- rstandard(model)
  par(mfrow = c(2, 2), cex.main = 0.9)
  plot(res, type = "o", pch = 16, cex = 0.5,
       xlab = "Time", ylab = "Standardised Residuals",
       main = paste(model_name, "— Residuals over Time"))
  abline(h =  0, col = "blue", lty = 2)
  abline(h =  3, col = "blue", lty = 3)
  abline(h = -3, col = "blue", lty = 3)
  hist(res, breaks = 20, xlab = "Standardised Residuals",
       main = paste(model_name, "— Histogram"),
       border = "white", freq = TRUE)
  curve(
    dnorm(x, mean(res, na.rm = TRUE), sd(res, na.rm = TRUE)) *
      length(res) * diff(hist(res, plot = FALSE, breaks = 20)$breaks)[1],
    add = TRUE, col = "blue", lwd = 2
  )
  qqnorm(res, main = paste(model_name, "— QQ Plot"), pch = 16, cex = 0.7)
  qqline(res, col = "blue", lty = 2, lwd = 1.5)
  acf(res, lag.max = lag.max,
      main = paste(model_name, "— ACF of Residuals"), ylab = "ACF")
  par(mfrow = c(1, 1))
  invisible(res)
}

# ── Function 10: tsdiag panel ─────────────────────────────────────────────────
plot_residuals_tsdiag <- function(model, lag.max = 40) {
  tsdiag(model, gof.lag = lag.max, omit.initial = FALSE)
}

# ── Function 11: Formal residual tests (Ljung-Box + Shapiro-Wilk) ─────────────
test_residuals <- function(model, model_name, lag.max = 40) {
  res <- rstandard(model)
  lb  <- Box.test(res,
                  lag  = min(lag.max, floor(length(res) / 5)),
                  type = "Ljung-Box")
  sw  <- shapiro.test(as.numeric(res))
  df <- data.frame(
    "Test"      = c("Ljung-Box", "Shapiro-Wilk"),
    "H0"        = c("No autocorrelation in residuals",
                    "Residuals are normally distributed"),
    "Statistic" = c(round(lb$statistic, 4), round(sw$statistic, 4)),
    "df"        = c(lb$parameter, NA),
    "p-value"   = c(round(lb$p.value, 4), round(sw$p.value, 4)),
    "Decision"  = c(
      ifelse(lb$p.value > 0.05,
             "Fail to Reject H0 → PASS", "Reject H0 → FAIL"),
      ifelse(sw$p.value > 0.05,
             "Fail to Reject H0 → PASS", "Reject H0 → FAIL")
    ),
    check.names = FALSE
  )
  list(lb = lb, sw = sw,
       table = knitr::kable(df,
                            caption = paste("Formal residual tests —", model_name),
                            align   = "llcccc"))
}

# ── Function 12: Benchmark forecasts (Naive, Drift, Mean) ─────────────────────
benchmark_forecasts <- function(y_train, y_test) {
  h <- length(y_test)
  list(
    Naive = forecast(naive(y_train, h = h)),
    Drift = rwf(y_train, drift = TRUE, h = h),
    Mean  = forecast(meanf(y_train, h = h))
  )
}

# ── Function 13: ETS forecast ─────────────────────────────────────────────────
ets_forecast <- function(y_train, h) {
  m <- ets(y_train)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 14: auto.arima forecast ──────────────────────────────────────────
auto_arima_forecast <- function(y_train, h, seasonal = FALSE) {
  m <- auto.arima(y_train, seasonal = seasonal,
                  stepwise = FALSE, approximation = FALSE)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 15: TBATS forecast ───────────────────────────────────────────────
tbats_forecast <- function(y_train, h) {
  m <- tbats(y_train)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 16: NNETAR forecast ──────────────────────────────────────────────
nnetar_forecast <- function(y_train, h, P = 1, repeats = 20) {
  m <- nnetar(y_train, P = P, repeats = repeats)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 17: STL decomposition forecast ───────────────────────────────────
stl_forecast <- function(y_train, h, s.window = 7) {
  m <- stlm(y_train, modelfunction = ar, s.window = s.window)
  list(model = m, forecast = forecast(m, h = h))
}

# ── Function 18: Forecast comparison table ────────────────────────────────────
compare_forecasts <- function(forecast_list, y_test,
                              caption = "Forecast Accuracy Comparison") {
  acc_df <- data.frame()
  for (name in names(forecast_list)) {
    fc_obj <- forecast_list[[name]]
    if (!is.null(fc_obj) && !is.null(fc_obj$mean)) {
      acc <- tryCatch(
        accuracy(fc_obj, y_test),
        error = function(e) NULL
      )
      if (is.null(acc)) next
      if (is.matrix(acc) && nrow(acc) >= 2) acc <- acc[2, ]
      temp_df <- data.frame(
        Method = name,
        RMSE   = round(as.numeric(acc["RMSE"]), 4),
        MAE    = round(as.numeric(acc["MAE"]),  4),
        MAPE   = round(as.numeric(acc["MAPE"]), 4),
        MASE   = round(as.numeric(acc["MASE"]), 4),
        stringsAsFactors = FALSE
      )
      acc_df <- rbind(acc_df, temp_df)
    }
  }
  acc_df       <- acc_df[order(acc_df$RMSE), ]
  acc_df$Rank  <- seq_len(nrow(acc_df))
  acc_df       <- acc_df[, c("Rank", "Method", "RMSE", "MAE", "MAPE", "MASE")]
  knitr::kable(acc_df, caption = caption, align = "clcccc",
               row.names = FALSE) %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}

# ── Function 19: Ensemble forecast ────────────────────────────────────────────
ensemble_forecast <- function(forecast_list, weights = NULL) {
  valid <- Filter(function(x) !is.null(x) && !is.null(x$mean), forecast_list)
  fc_matrix <- do.call(cbind, lapply(valid, function(x) as.numeric(x$mean)))
  if (is.null(weights)) weights <- rep(1 / ncol(fc_matrix), ncol(fc_matrix))
  ens_mean <- as.numeric(fc_matrix %*% weights)
  ts(ens_mean,
     start     = start(valid[[1]]$mean),
     frequency = frequency(valid[[1]]$mean))
}

# ── Function 20: Forecast plot (single model with PIs) ────────────────────────
plot_forecast_simple <- function(fc, y_train, y_test, model_name) {
  h <- length(y_test)

  # Cast everything to plain numeric — prevents ts class conflict in plot()
  x_all <- c(as.numeric(time(y_train)), as.numeric(time(y_test)))
  y_all <- c(as.numeric(y_train),       as.numeric(y_test))

  plot(x_all, y_all, type = "n",
       main = paste(model_name, "— Forecast, Next", h, "Days"),
       ylab = "Price (AUD)", xlab = "Time")

  if (!is.null(fc$lower) && !is.null(fc$upper)) {
    fc_time <- as.numeric(time(fc$mean))

    # 95% PI (wider, lighter)
    polygon(c(fc_time, rev(fc_time)),
            c(as.numeric(fc$lower[, "95%"]),
              rev(as.numeric(fc$upper[, "95%"]))),
            col = adjustcolor("steelblue", alpha.f = 0.2), border = NA)

    # 80% PI (narrower, darker)
    polygon(c(fc_time, rev(fc_time)),
            c(as.numeric(fc$lower[, "80%"]),
              rev(as.numeric(fc$upper[, "80%"]))),
            col = adjustcolor("steelblue", alpha.f = 0.4), border = NA)
  }

  lines(as.numeric(time(y_train)), as.numeric(y_train),
        col = "black",      lwd = 1.5)
  lines(as.numeric(time(fc$mean)), as.numeric(fc$mean),
        col = "steelblue4", lwd = 2)
  lines(as.numeric(time(y_test)),  as.numeric(y_test),
        col = "red",        lwd = 2)

  legend("topleft",
         legend = c("Training", "Forecast", "80% PI", "95% PI", "Actual"),
         col    = c("black", "steelblue4",
                    adjustcolor("steelblue", 0.4),
                    adjustcolor("steelblue", 0.2), "red"),
         lwd    = c(1.5, 2, 8, 8, 2), bty = "n", cex = 0.7)
}


# ── Function 21: Multi-method forecast comparison plot ────────────────────────
plot_forecast_comparison <- function(forecast_list, y_train, y_test,
                                     title = "Forecast Comparison") {
  all_fc_vals <- unlist(lapply(forecast_list,
                               function(x) if (!is.null(x$mean)) as.numeric(x$mean)))
  ylim_range  <- range(c(as.numeric(y_train), as.numeric(y_test),
                          all_fc_vals), na.rm = TRUE)

  # Cast to plain numeric — prevents ts class conflict in plot()
  x_all <- c(as.numeric(time(y_train)), as.numeric(time(y_test)))
  y_all <- c(as.numeric(y_train),       as.numeric(y_test))

  plot(x_all, y_all, type = "n",
       xlab = "Time", ylab = "Price (AUD)",
       main = title, ylim = ylim_range)
  grid(col = "lightgray")

  lines(as.numeric(time(y_train)), as.numeric(y_train),
        col = "black", lwd = 2)
  lines(as.numeric(time(y_test)),  as.numeric(y_test),
        col = "red",   lwd = 2, lty = 2)

  pal <- c("blue", "darkgreen", "purple", "darkorange",
           "brown", "deeppink", "cyan4", "navy", "tomato")

  for (i in seq_along(forecast_list)) {
    fc <- forecast_list[[i]]
    if (!is.null(fc) && !is.null(fc$mean)) {
      lines(as.numeric(time(fc$mean)), as.numeric(fc$mean),
            col = pal[(i - 1) %% length(pal) + 1],
            lwd = 1.5, lty = 1)
    }
  }

  legend("topleft",
         legend = c("Training", "Actual (test)",
                    names(forecast_list)),
         col    = c("black", "red", pal[seq_along(forecast_list)]),
         lty    = c(1, 2, rep(1, length(forecast_list))),
         lwd    = 2, cex = 0.6, bty = "n")
}
```

# Introduction

This study applies a systematic ARIMA time series modelling workflow to the daily adjusted closing price of Commonwealth Bank of Australia (CBA.AX) — one of Australia's "Big Four" banks and the largest company by market capitalisation on the ASX 200. CBA's price movements are closely watched as a barometer of the broader Australian financial sector, domestic housing market conditions, and the Reserve Bank of Australia's (RBA) monetary policy trajectory.

The analysis follows the Box–Jenkins methodology (Cryer & Chan, 2008) as taught in MATH1318/2204, structured across six stages:

1.  **Exploratory Data Analysis** — visual inspection of trend, variance, and distributional properties.
2.  **Transformation** — Box-Cox variance stabilisation where warranted.
3.  **Differencing & Stationarity** — achieving a time-invariant mean via differencing, verified by formal unit-root tests.
4.  **Model Specification** — AR and MA order identification via ACF, PACF, EACF, and BIC tables.
5.  **Model Fitting & Residual Diagnostics** — CSS and ML estimation with full Module 7 residual analysis and overparameterised model checks.
6.  **Forecasting** — producing point forecasts and prediction intervals using multiple methods.

# Data Acquisition

## Define ASX Top 10 Tickers

```{r asx10-tickers}
#| label: asx10-tickers
#| cache: true

asx10_tickers <- tibble(
  symbol  = c("BHP",  "CBA",  "CSL",  "NAB",  "WBC",
               "ANZ",  "WES",  "MQG",  "RIO",  "TLS"),
  company = c("BHP Group",               "Commonwealth Bank",
               "CSL Limited",             "National Australia Bank",
               "Westpac Banking",         "ANZ Banking Group",
               "Wesfarmers",              "Macquarie Group",
               "Rio Tinto",               "Telstra"),
  sector  = c("Materials",   "Financials",  "Healthcare",
               "Financials",  "Financials",  "Financials",
               "Consumer",    "Financials",  "Materials",  "Telecom")
) %>%
  mutate(symbol_yf = paste0(symbol, ".AX"))

cat("Tickers defined:", nrow(asx10_tickers), "\n")
print(asx10_tickers[, c("symbol", "company", "sector", "symbol_yf")])
```

## Download Daily Price History

```{r asx10-prices}
#| label: asx10-prices
#| cache: true

safe_tq_get <- possibly(tq_get, otherwise = NULL)

asx10_prices_raw <- asx10_tickers$symbol_yf %>%
  map_dfr(~ safe_tq_get(.x, get = "stock.prices",
                         from = Sys.Date() - 730, to = Sys.Date()))

cat("Total rows downloaded:", nrow(asx10_prices_raw), "\n")
cat("Unique tickers retrieved:", n_distinct(asx10_prices_raw$symbol), "\n")
```

## Construct Time Series Objects

```{r cba-ts-setup}
#| label: cba-ts-setup
#| cache: true

cba_data <- asx10_prices_raw %>%
  filter(symbol == "CBA.AX") %>%
  arrange(date)

cba_prices_ts <- ts(cba_data$adjusted, start = c(2024, 1), frequency = 252)
cba_volumes_ts <- ts(cba_data$volume,   start = c(2024, 1), frequency = 252)

n_train <- length(cba_prices_ts) - 90
y_train <- window(cba_prices_ts, end   = time(cba_prices_ts)[n_train])
y_test  <- window(cba_prices_ts, start = time(cba_prices_ts)[n_train + 1])

cat("CBA observations (total):", length(cba_prices_ts), "\n")
cat("Training observations:   ", length(y_train), "\n")
cat("Test observations:       ", length(y_test),  "\n")
cat("Date range:", format(min(cba_data$date), "%d %b %Y"),
    "to",         format(max(cba_data$date), "%d %b %Y"), "\n")
cat("Price range: AUD", round(min(cba_data$adjusted), 2),
    "–",               round(max(cba_data$adjusted), 2), "\n")
```

# Exploratory Data Analysis (EDA)

## Stationarity Inspection

### Visual Inspection

```{r fig-eda-raw-ts}
#| label: fig-eda-raw-ts
#| fig-cap: "CBA adjusted closing price — training sample (April 2024 onward). The sustained upward quadratic trend, the sharp level drop near 2025.5, and the widening amplitude of fluctuations as the price level rises."
#| fig-height: 4

plot_ts(y_train, main = "CBA Adjusted Closing Price — Training Sample",
        ylab = "Price (AUD)", add_mean = TRUE)
```

```{r fig-eda-acf-pacf-raw}
#| label: fig-eda-acf-pacf-raw
#| fig-cap: "ACF (left) and PACF (right) of the raw CBA price series. The ACF's slow linear decay and the PACF's single dominant spike at lag 1, confirming non-stationarity before any statistical testing."
#| fig-height: 4

plot_acf_pacf(y_train,
              acf_title  = "ACF — CBA Raw Price",
              pacf_title = "PACF — CBA Raw Price")
```

### Statistical Tests

```{r stat-eda-stationarity}
#| label: stat-eda-stationarity

stat_raw <- check_stationarity(y_train,
  "Stationarity Tests — Raw CBA Price (Training Sample)")
stat_raw$table
```

### Interpretation

**Trend.** @fig-eda-raw-ts shows a sustained upward quadratic trend from approximately AUD 110 at the start of 2024 to a peak near AUD 185 by mid-2025. The horizontal mean line sits well below the peak of the series, confirming the absence of a stable central tendency — a necessary condition for ARIMA modelling is violated from the outset.

**Seasonality.** There is no evidence of a fixed-frequency repeating cycle within this two-year window. Price fluctuations appear stochastic rather than deterministic, suggesting that a non-seasonal ARIMA specification is appropriate. Any apparent periodicity aligns with irregular macroeconomic events (RBA meeting cycles, CBA earnings seasons) rather than a strict calendar pattern.

**Heteroscedasticity.** The magnitude of daily fluctuations increases materially as the price level rises, particularly after 2025.0. This proportional relationship between price level and volatility — a characteristic of geometric Brownian motion — suggests that the raw series will require variance-stabilising investigation before ARIMA modelling.

**Change Point.** A significant level shift occurs shortly after 2025.5, where the price drops from roughly AUD 175 to AUD 150 in a very short interval — likely reflecting a macro shock such as an unexpected RBA policy decision, changes to bank capital requirements, or a global credit event — followed by elevated volatility and partial recovery.

**ACF — slow linear decay** (@fig-eda-acf-pacf-raw, left). The autocorrelations remain large and positive across many lags, declining so slowly that they barely enter the significance bands even at lag 40. This persistence confirms the series has very long memory: each day's price level carries the weight of all preceding observations forward indefinitely.

**PACF — single dominant spike at lag 1** (@fig-eda-acf-pacf-raw, right). Once the chain of correlations is removed, only the immediately preceding observation retains direct predictive power. This is the canonical PACF signature of a unit-root process.

**Stationarity tests.** The ADF (p = `r round(stat_raw$adf$p.value, 4)`) and PP (p = `r round(stat_raw$pp$p.value, 4)`) tests both fail to reject the null hypothesis of a unit root, while KPSS (p = `r round(stat_raw$kpss$p.value, 4)`) rejects its stationarity null. All three tests agree: the series contains a stochastic trend that must be removed before modelling.

------------------------------------------------------------------------

## Normality & Distributional Inspection

### Visual Inspection

```{r fig-eda-norm}
#| label: fig-eda-norm
#| fig-cap: "QQ plot (left) and histogram with normal overlay (right) for the raw CBA training price series. The S-shaped deviation from the reference line and a non-symmetric histogram reject the assumption that the data follow a normal distribution."
#| fig-height: 4

plot_norm(y_train, qq_title = "QQ Plot — Raw CBA Price",
          hist_title = "Histogram — Raw CBA Price", xlab = "Price (AUD)")
```

### Statistical Tests

```{r stat-eda-norm-summary}
#| label: stat-eda-norm-summary

Summary <- c(
  Mean   = round(mean(y_train),               2),
  Median = round(median(y_train),             2),
  SD     = round(sd(y_train),                 2),
  Min    = round(min(y_train),                2),
  Max    = round(max(y_train),                2),
  Range  = round(max(y_train) - min(y_train), 2)
)
knitr::kable(t(as.data.frame(Summary)),
  caption = "Descriptive statistics — CBA.AX adjusted closing price (training sample).",
  align = "c")
```

```{r stat-eda-normality}
#| label: stat-eda-normality

normality_raw <- check_normality(y_train, "Shapiro-Wilk Test — Raw CBA Price")
normality_raw$table
```

### Interpretation

**Central tendency.** The mean (AUD `r round(mean(y_train), 2)`) and median (AUD `r round(median(y_train), 2)`) are closely aligned, but this proximity is misleading: it results from the series spending roughly equal time in its lower and upper price ranges, not from any genuine normality.

**Dispersion.** The standard deviation of AUD `r round(sd(y_train), 2)` reflects the wide price range traversed over the training window — a direct consequence of the sustained upward trend in @fig-eda-raw-ts rather than a meaningful measure of single-period volatility.

**QQ plot** (@fig-eda-norm, left). The S-shaped deviation from the theoretical line at both tails indicates lighter tails than a normal distribution — a natural consequence of the bounded lower tail (prices cannot fall below zero) and the right-skew imposed by the trend.

**Histogram** (@fig-eda-norm, right). The multimodal profile with at least two visible clusters corresponds to the lower pre-2025 and higher 2025-peak price ranges. No single normal density can describe this bimodal structure, which is a direct symptom of the non-stationarity identified above.

**Shapiro-Wilk test.** With W = `r round(normality_raw$result$statistic, 4)` and p ≈ `r format(normality_raw$result$p.value, scientific = TRUE, digits = 2)`, we decisively reject normality. Transformation and differencing are required before valid ARIMA estimation.

------------------------------------------------------------------------

## Synthesis of EDA Findings

**Variance stabilisation required.** The Shapiro-Wilk test (p ≈ 0), S-shaped QQ deviations, and visually apparent heteroscedasticity in @fig-eda-raw-ts collectively confirm non-constant variance proportional to price level. A Box-Cox transformation will be evaluated in @sec-transformation.

**Mean stabilisation required.** The unanimous agreement of ADF (p = `r round(stat_raw$adf$p.value, 4)`), PP (p = `r round(stat_raw$pp$p.value, 4)`), and KPSS (p = `r round(stat_raw$kpss$p.value, 4)`) confirms the presence of a unit root. The slow ACF decay in @fig-eda-acf-pacf-raw provides visual corroboration. First-order differencing (d = 1) will be applied in @sec-stationarity.

# Transformation {#sec-transformation}

## Box-Cox Profile Analysis

### Visual Inspection

```{r fig-boxcox-profile}
#| label: fig-boxcox-profile
#| fig-cap: "The Box-Cox log-likelihood profile for the CBA training series indicates an optimal λ ≈ −0.3."
#| fig-height: 5

BC <- BoxCox.ar(y = y_train, lambda = seq(-2, 2, 0.01))
```

### Statistical Tests

```{r stat-boxcox-ci}
#| label: stat-boxcox-ci

lambda_opt   <- BC$lambda[which.max(BC$loglike)]
lambda_lower <- BC$ci[1]
lambda_upper <- BC$ci[2]

bc_df <- data.frame(
  Parameter = c("Optimal λ", "95% CI Lower", "95% CI Upper",
                "CI includes λ = 0 (log)?", "CI includes λ = 1 (none)?"),
  Value     = c(round(lambda_opt,   4), round(lambda_lower, 4),
                round(lambda_upper, 4),
                ifelse(lambda_lower <= 0 & lambda_upper >= 0, "YES", "NO"),
                ifelse(lambda_lower <= 1 & lambda_upper >= 1, "YES", "NO"))
)
knitr::kable(bc_df, caption = "Box-Cox parameter estimation results.",
             align = "lc", col.names = c("Parameter", "Value"))
```

### Interpretation

The optimal λ (`r round(lambda_opt, 4)`) sits close to zero, and the 95% CI from @fig-boxcox-profile determines whether a log transformation is statistically supported. For CBA daily prices, the CI typically includes λ = 0 and excludes λ = 1 — providing formal support for the log transform and aligning with financial econometric theory: log-prices follow a random walk with drift, converting multiplicative price movements into additive log-returns.

------------------------------------------------------------------------

## Applied Transformation

### Visual Inspection

```{r fig-bc-compare}
#| label: fig-bc-compare
#| fig-cap: "Original CBA price series (top) and Box-Cox transformed series (bottom). A successful transformation would show reduced heteroscedasticity — specifically, fluctuation amplitude becoming more uniform across the sample."
#| fig-height: 6

BC_train <- (y_train^lambda_opt - 1) / lambda_opt
BC_test  <- (y_test^lambda_opt  - 1) / lambda_opt

par(mfrow = c(2, 1), cex = 0.85)
plot_ts(y_train,  main = "CBA — Original Series",          ylab = "Price (AUD)")
plot_ts(BC_train, main = "CBA — Box-Cox Transformed Series", ylab = "BC(Price)")
par(mfrow = c(1, 1))
```

```{r fig-bc-acf-pacf}
#| label: fig-bc-acf-pacf
#| fig-cap: "ACF (left) and PACF (right) of the Box-Cox transformed series. The slow ACF decay and single dominant PACF spike at lag 1 mirror the patterns in @fig-eda-acf-pacf-raw, confirming the transformation has not altered the fundamental non-stationary structure."
#| fig-height: 4

plot_acf_pacf(BC_train,
              acf_title  = "ACF — Box-Cox Transformed Series",
              pacf_title = "PACF — Box-Cox Transformed Series")
```

```{r fig-bc-norm}
#| label: fig-bc-norm
#| fig-cap: "QQ plot (left) and histogram with normal overlay (right) for the Box-Cox transformed series. Near-identical tail behaviour to the original series (@fig-eda-norm) indicates that the transformation has not materially improved distributional normality."
#| fig-height: 4

plot_norm(BC_train, qq_title = "QQ Plot — BC Transformed Series",
          hist_title = "Histogram — BC Transformed Series", xlab = "BC(Price)")
```

### Statistical Tests

```{r stat-bc-normality}
#| label: stat-bc-normality

normality_bc <- check_normality(BC_train,
  "Shapiro-Wilk Test — Box-Cox Transformed Series")
normality_bc$table
```

```{r stat-bc-stationarity}
#| label: stat-bc-stationarity

stat_bc <- check_stationarity(BC_train, "Stationarity Tests — Box-Cox Transformed Series")
stat_bc$table
```

### Interpretation

**Time series plot** (@fig-bc-compare). The Box-Cox transformed series retains the same overall shape — the upward trend persists and the amplitude of fluctuations continues to grow after 2025.0. The transformation has not materially altered the non-stationary structure or visibly reduced heteroscedasticity, consistent with financial series where variance is driven primarily by the stochastic trend.

**ACF/PACF** (@fig-bc-acf-pacf). The slow linear ACF decay and single dominant PACF spike at lag 1 are nearly identical to the raw series in @fig-eda-acf-pacf-raw. Non-stationarity in the mean remains unresolved.

**QQ plot and histogram** (@fig-bc-norm). The S-shaped QQ deviation and multimodal histogram are essentially indistinguishable from @fig-eda-norm. No meaningful distributional improvement has been achieved.

**Normality.** W = `r round(normality_bc$result$statistic, 4)`, p = `r format(normality_bc$result$p.value, scientific = TRUE, digits = 2)` — marginally worse than the original series. Normality null is again decisively rejected.

**Stationarity.** ADF (p = `r round(stat_bc$adf$p.value, 4)`), PP (p = `r round(stat_bc$pp$p.value, 4)`), and KPSS (p = `r round(stat_bc$kpss$p.value, 4)`) confirm the transformed series remains non-stationary.

------------------------------------------------------------------------

## Decision: Proceed Without Transformation

The Box-Cox transformation produces no substantive improvement in normality, variance stability, or stationarity. In accordance with the principle of parsimony (Cryer & Chan, 2008), the transformation is rejected. **The analysis proceeds on the original adjusted price series.** The next phase applies first-order differencing (d = 1) to stabilise the mean.

# Addressing Non-Stationarity {#sec-stationarity}

## First Differencing

### Visual Inspection

```{r fig-diff-series}
#| label: fig-diff-series
#| fig-cap: "Original CBA price series (top) and first-differenced series (bottom). The differenced series fluctuates around zero (dashed blue line), confirming the stochastic trend has been successfully removed. Large spikes near 2025.5 correspond to high-impact trading days."
#| fig-height: 6

diff_y_train <- diff(y_train, differences = 1)

par(mfrow = c(2, 1), cex = 0.85)
plot_ts(y_train, main = "(a) Original CBA Price", ylab = "Price (AUD)")
plot_ts(diff_y_train, main = "(b) First-Differenced Series (d = 1)",
        ylab = "Δ Price (AUD)", add_mean = TRUE)
par(mfrow = c(1, 1))
```

```{r fig-diff-acf-pacf}
#| label: fig-diff-acf-pacf
#| fig-cap: "ACF (left) and PACF (right) of the first-differenced CBA series. The immediate drop-off with most bars within the 95% confidence bands confirms stationarity and reveals the underlying ARMA structure for model specification."
#| fig-height: 4

plot_acf_pacf(diff_y_train,
              acf_title  = "ACF — Δ CBA Price (d = 1)",
              pacf_title = "PACF — Δ CBA Price (d = 1)")
```

### Statistical Tests

```{r stat-diff-stationarity}
#| label: stat-diff-stationarity

stat_d1 <- check_stationarity(diff_y_train,
  "Stationarity Tests — First-Differenced Series (d = 1)")
stat_d1$table
```

### Interpretation

**Trend removal** (@fig-diff-series). The upward trend in panel (a) is completely removed in panel (b). The differenced series fluctuates around a stable horizontal mean of approximately zero, confirming that first-order differencing was sufficient; a second difference is not required and would risk over-differencing.

**Volatility and outliers.** @fig-diff-series panel (b) highlights elevated volatility near 2025.5 corresponding to the level-shift event from the EDA. These transient volatility clusters are characteristic of bank equity daily returns and reflect high-impact events — RBA cash rate decisions, APRA regulatory announcements, global credit risk repricing. The ARIMA framework models the autocorrelative structure; persistent volatility clustering would require a GARCH extension.

**ACF/PACF** (@fig-diff-acf-pacf). Both plots drop off quickly after initial lags with the majority of bars within the 95% bands. The isolated significant spike at lag 5 in both ACF and PACF represents the residual autocorrelative signal that model specification will investigate.

**Stationarity tests.** ADF (p = `r round(stat_d1$adf$p.value, 4)`) and PP (p = `r round(stat_d1$pp$p.value, 4)`) both reject the unit-root null; KPSS (p = `r round(stat_d1$kpss$p.value, 4)`) fails to reject its stationarity null. All three tests unanimously confirm stationarity. **The integration order is confirmed as d = 1.**

**ACF for MA order (q) identification.** The isolated spike at lag 5 with no significant autocorrelation at lower lags is an atypical pattern. It may indicate a five-day settlement cycle or weekly rebalancing effect, or transient noise — which the EACF and BIC analysis will help distinguish.

**PACF for AR order (p) identification.** Significant PACF spikes at lag 5 (and lag 8) suggest potential AR influence at those intervals. An ARIMA(5,1,0) is implied by strict PACF reading, though parsimony and BIC may favour simpler alternatives.

------------------------------------------------------------------------

## Decision: Proceed With First-Differenced Original Series

Based on the evidence above, the analysis proceeds with the **original (untransformed) series at d = 1**. Further differencing (d = 2) is rejected for three reasons:

1.  **Risk of over-differencing.** d = 1 has already achieved statistically confirmed stationarity across all three tests. A second difference introduces artificial moving-average structure and inflates residual variance unnecessarily.
2.  **Loss of interpretability.** d = 1 transforms prices into daily changes (returns) — a natural, meaningful financial quantity. d = 2 would model the "change in changes" (acceleration), which has no natural financial interpretation and is highly sensitive to noise.
3.  **Parsimony.** A stationary d = 1 series allows for a simpler ARIMA structure less likely to overfit, preserving out-of-sample generalisation on the 90-day test set.

# Model Specification

## Order Identification from ACF and PACF

The standard rules (Cryer & Chan, 2008) for translating @fig-diff-acf-pacf into ARMA orders are:

-   ACF cuts off at lag q, PACF decays geometrically → **MA(q)**
-   PACF cuts off at lag p, ACF decays geometrically → **AR(p)**
-   Both decay geometrically → **mixed ARMA(p,q)**

Following these rules, the first-differenced series exhibits predominantly white-noise behaviour in early lags (1–4 and 6–11), with no persistent autocorrelation structure. Isolated spikes at lag 5 and lag 8 are treated as potential signals but evaluated against lower-order models through EACF and BIC — attempting to model lag 5 or lag 8 directly through standard AR/MA orders risks over-parameterisation. The analysis prioritises parsimonious low-order candidates consistent with the near-random-walk nature of daily equity returns.

## Extended Autocorrelation Function (EACF) {#sec-eacf}

```{r stat-eacf}
#| label: stat-eacf

cat("=== EACF — First-Differenced CBA Price ===\n")
eacf(diff_y_train, ar.max = 5, ma.max = 5)
```

The 'o' symbols mark the upper-left corner of a white-noise triangle, identifying candidate ARMA(p,q) pairs. The EACF vertex at AR = 0, MA = 1 identifies ARIMA(0,1,1) as the most parsimonious candidate; a secondary 'o' at (1,1) suggests ARIMA(1,1,1) as a viable mixed alternative. The EACF effectively ignores the isolated high-lag spikes from @fig-diff-acf-pacf, reinforcing the parsimony argument.

## BIC Subset Selection {#sec-bic}

### 10×10 Search Space

```{r fig-bic-10x10-subsets}
#| label: fig-bic-10x10-subsets
#| fig-cap: "BIC subset selection — 10×10 search space. The top-ranked row (lowest BIC) identifies the most parsimonious model across a broad AR/MA order grid. Shaded cells show retained parameters at each order level."
#| fig-height: 5

res_bic_10 <- armasubsets(y = diff_y_train, nar = 10, nma = 10,
                          y.name = "p", ar.method = "ols")
plot(res_bic_10)
```

The 10×10 BIC search in @fig-bic-10x10-subsets ranks models across a broad grid. The lowest-BIC row identifies ARIMA(7,1,0) — where only the 7th-order AR component minimises information loss — potentially capturing a weekly cyclicality or delayed market reaction specific to CBA. The second-ranked row suggests a more complex higher-order specification; the substantial BIC penalty increase from row 1 to row 2 signals diminishing returns from additional parameters.

### 5×5 Search Space

```{r fig-bic-5x5-subsets}
#| label: fig-bic-5x5-subsets
#| fig-cap: "BIC subset selection — 5×5 search space. The restricted grid provides higher resolution for short-lag dependencies. The top row identifies ARIMA(3,1,0) as optimal within this search space."
#| fig-height: 5

res_bic_5 <- armasubsets(y = diff_y_train, nar = 5, nma = 5,
                         y.name = "p", ar.method = "ols")
plot(res_bic_5)
```

Within the 5×5 restricted grid (@fig-bic-5x5-subsets), ARIMA(3,1,0) achieves the lowest BIC — the shading at p-lag3 suggests a three-day autoregressive structure capturing short-term momentum or mean reversion. The second-ranked specification incurs a BIC increase of approximately 6 units, which constitutes strong statistical evidence in favour of the simpler model. ARIMA(1,1,3) is retained as a secondary BIC candidate despite this penalty.

## Candidate Model Selection

```{r stat-candidates}
#| label: stat-candidates

candidates_df <- data.frame(
  "Candidate Model"      = c("ARIMA(0,1,1)", "ARIMA(1,1,0)", "ARIMA(1,1,1)",
                              "ARIMA(3,1,0)", "ARIMA(1,1,3)",
                              "ARIMA(7,1,0)", "ARIMA(7,1,9)"),
  "Specification Source" = c("EACF", "EACF", "EACF / ACF & PACF",
                              "BIC 5×5", "BIC 5×5",
                              "BIC 10×10", "BIC 10×10"),
  "Rationale" = c(
    "MA(1) captures the one-period impact of short-term shocks — e.g., an RBA rate change absorbed within a single trading day.",
    "AR(1) models short-term momentum: today's return as a partial function of yesterday's, reflecting institutional buying or selling pressure.",
    "Mixed ARMA(1,1) captures both momentum (AR) and shock persistence (MA) — the most complete low-order specification.",
    "Three-lag AR captures momentum over approximately one trading week, reflecting short-term price trends.",
    "MA(3) captures shock persistence over three trading days, implying market surprises influence prices for nearly a full week.",
    "7-lag AR captures potential weekly cyclicality (≈ 7 trading days), detecting systematic day-of-week effects.",
    "High-order mixed model with 7 AR lags (≈ 2 trading weeks) and 9 MA lags — maximum flexibility at high parameterisation cost."
  ),
  check.names = FALSE
)

kbl(candidates_df,
    caption  = "Final candidate ARIMA models for the CBA daily price series.",
    booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE, width = "3.5cm") %>%
  column_spec(2, width = "4cm") %>%
  column_spec(3, width = "8.5cm")
```

The seven candidates span a spectrum from parsimonious low-order models (ARIMA(0,1,1), ARIMA(1,1,0)) to high-order BIC-selected specifications (ARIMA(7,1,0), ARIMA(7,1,9)). This deliberate range tests whether added complexity resolves genuine structure or merely absorbs noise — a distinction that residual diagnostics and information criteria will determine.

# Model Fitting

All models are fitted to the **raw (untransformed) original price series** using `Arima()`. Both CSS and ML estimation are applied. CSS provides a computationally efficient approximation; ML maximises the exact likelihood. In a sample of this size (n ≈ 420), the two methods should converge to nearly identical estimates — any material divergence is a signal of mis-specification or near-boundary parameters. `include.mean = FALSE` constrains the drift to zero, consistent with the no-drift assumption for CBA daily price changes.

```{r fit-models}
#| label: fit-models

# ── Fit all 7 candidates under CSS and ML ─────────────────────────────────────
model_011_css <- Arima(y_train, order = c(0,1,1), method = "CSS")
model_011_ml  <- Arima(y_train, order = c(0,1,1), method = "ML")

model_110_css <- Arima(y_train, order = c(1,1,0), method = "CSS")
model_110_ml  <- Arima(y_train, order = c(1,1,0), method = "ML")

model_111_css <- Arima(y_train, order = c(1,1,1), method = "CSS")
model_111_ml  <- Arima(y_train, order = c(1,1,1), method = "ML")

model_310_css <- Arima(y_train, order = c(3,1,0), method = "CSS")
model_310_ml  <- Arima(y_train, order = c(3,1,0), method = "ML")

model_113_css <- Arima(y_train, order = c(1,1,3), method = "CSS")
model_113_ml  <- Arima(y_train, order = c(1,1,3), method = "ML")

model_710_css <- Arima(y_train, order = c(7,1,0), method = "CSS")
model_710_ml  <- Arima(y_train, order = c(7,1,0), method = "ML")

model_719_css <- Arima(y_train, order = c(7,1,9), method = "CSS")
model_719_ml  <- Arima(y_train, order = c(7,1,9), method = "ML")

cat("All 7 candidate models fitted successfully under CSS and ML.\n")
```

## ARIMA(0,1,1)

```{r stat-coef-011}
#| label: stat-coef-011

make_coef_table(model_011_css, "ARIMA(0,1,1) — CSS Estimation")
make_coef_table(model_011_ml,  "ARIMA(0,1,1) — ML Estimation")
```

```{r fig-fitted-011}
#| label: fig-fitted-011
#| fig-cap: "ARIMA(0,1,1) fitted values overlaid on the original CBA price series under CSS (left) and ML (right). Near-identical trajectories indicate a numerically stable, well-identified model."
#| fig-height: 4

plot_css_ml(model_011_css, model_011_ml, y_train, spec = "ARIMA(0,1,1)")
```

The MA(1) coefficient θ₁ captures the one-period lingering effect of a price shock — for example, a surprise RBA announcement that initiates a price adjustment which fades within one trading day. For the model to be well-identified, θ₁ must be statistically significant (p \< 0.05) under both CSS and ML, and must satisfy the invertibility condition \|θ₁\| \< 1. If the MA(1) term is non-significant, this implies CBA daily price changes carry no exploitable one-day shock memory — consistent with a pure random walk (ARIMA(0,1,0)). As shown in @fig-fitted-011, near-identical CSS and ML fitted trajectories confirm numerical stability regardless of coefficient significance.

## ARIMA(1,1,0)

```{r stat-coef-110}
#| label: stat-coef-110

make_coef_table(model_110_css, "ARIMA(1,1,0) — CSS Estimation")
make_coef_table(model_110_ml,  "ARIMA(1,1,0) — ML Estimation")
```

```{r fig-fitted-110}
#| label: fig-fitted-110
#| fig-cap: "ARIMA(1,1,0) fitted values overlaid on the original CBA price series under CSS (left) and ML (right)."
#| fig-height: 4

plot_css_ml(model_110_css, model_110_ml, y_train, spec = "ARIMA(1,1,0)")
```

The AR(1) coefficient φ₁ models yesterday's price change as a direct predictor of today's. A positive φ₁ implies short-term momentum — a day of above-average CBA returns tends to be followed by another, consistent with institutional fund-flow inertia in large-cap ASX bank equities. A negative φ₁ implies mean reversion — sharp moves tend to partially retrace the following day. Agreement between CSS and ML estimates in @fig-fitted-110 signals a stable specification.

## ARIMA(1,1,1)

```{r stat-coef-111}
#| label: stat-coef-111

make_coef_table(model_111_css, "ARIMA(1,1,1) — CSS Estimation")
make_coef_table(model_111_ml,  "ARIMA(1,1,1) — ML Estimation")
```

```{r fig-fitted-111}
#| label: fig-fitted-111
#| fig-cap: "ARIMA(1,1,1) fitted values overlaid on the original CBA price series under CSS (left) and ML (right)."
#| fig-height: 4

plot_css_ml(model_111_css, model_111_ml, y_train, spec = "ARIMA(1,1,1)")
```

The combined AR(1)+MA(1) specification captures both momentum and shock persistence but risks AR–MA parameter cancellation — where the AR and MA roots partially offset each other, inflating standard errors and rendering one or both terms non-significant. If ML finds either the AR(1) or MA(1) insignificant while the simpler models produce significant individual estimates, the principle of parsimony dictates selecting the simpler specification. @fig-fitted-111 provides the visual CSS-versus-ML alignment check.

## ARIMA(3,1,0)

```{r stat-coef-310}
#| label: stat-coef-310

make_coef_table(model_310_css, "ARIMA(3,1,0) — CSS Estimation")
make_coef_table(model_310_ml,  "ARIMA(3,1,0) — ML Estimation")
```

```{r fig-fitted-310}
#| label: fig-fitted-310
#| fig-cap: "ARIMA(3,1,0) fitted values overlaid on the original CBA price series under CSS (left) and ML (right)."
#| fig-height: 4

plot_css_ml(model_310_css, model_310_ml, y_train, spec = "ARIMA(3,1,0)")
```

The three-lag AR structure allows today's price change to depend on changes over the past three trading days — approximately one calendar week of market information. This may capture a slow-decay momentum effect where institutional orders are executed over several days, or a partial mean-reversion following a multi-day directional run. However, the significance of all three AR coefficients under ML must be jointly assessed: if AR(2) and AR(3) are non-significant, the model collapses toward ARIMA(1,1,0) and the extra parameters are not justified.

## ARIMA(1,1,3)

```{r stat-coef-113}
#| label: stat-coef-113

make_coef_table(model_113_css, "ARIMA(1,1,3) — CSS Estimation")
make_coef_table(model_113_ml,  "ARIMA(1,1,3) — ML Estimation")
```

```{r fig-fitted-113}
#| label: fig-fitted-113
#| fig-cap: "ARIMA(1,1,3) fitted values overlaid on the original CBA price series under CSS (left) and ML (right)."
#| fig-height: 4

plot_css_ml(model_113_css, model_113_ml, y_train, spec = "ARIMA(1,1,3)")
```

The MA(3) component alongside AR(1) models shock persistence over three trading days — implying that a surprise market event influences CBA prices for nearly a full trading week after it occurs. With four estimated parameters, this model carries moderate over-parameterisation risk. CSS–ML divergence in any of the MA terms visible in @fig-fitted-113 would indicate numerical instability in the likelihood surface, a common symptom of over-specified ARMA models on near-random-walk series.

## ARIMA(7,1,0)

```{r stat-coef-710}
#| label: stat-coef-710

make_coef_table(model_710_css, "ARIMA(7,1,0) — CSS Estimation")
make_coef_table(model_710_ml,  "ARIMA(7,1,0) — ML Estimation")
```

```{r fig-fitted-710}
#| label: fig-fitted-710
#| fig-cap: "ARIMA(7,1,0) fitted values overlaid on the original CBA price series under CSS (left) and ML (right). With seven AR parameters, any CSS–ML divergence is a strong indicator of over-fitting."
#| fig-height: 4

plot_css_ml(model_710_css, model_710_ml, y_train, spec = "ARIMA(7,1,0)")
```

The 7-lag AR structure was identified as the BIC-optimal specification in the 10×10 search, suggesting a potential weekly cyclicality in CBA price changes — consistent with systematic day-of-week effects documented in bank equity trading (e.g., lower liquidity on Mondays, institutional rebalancing on Fridays). However, seven parameters impose a substantial information penalty. For this model to be justified, a majority of AR coefficients must be statistically significant under ML; a sparse pattern where only AR(7) is significant while AR(1)–AR(6) are not would suggest the BIC is detecting a specific seasonal pulse rather than a genuine AR structure.

## ARIMA(7,1,9)

```{r stat-coef-719}
#| label: stat-coef-719

make_coef_table(model_719_css, "ARIMA(7,1,9) — CSS Estimation")
make_coef_table(model_719_ml,  "ARIMA(7,1,9) — ML Estimation")
```

```{r fig-fitted-719}
#| label: fig-fitted-719
#| fig-cap: "ARIMA(7,1,9) fitted values overlaid on the original CBA price series under CSS (left) and ML (right). With 16 estimated parameters, this model is the most likely to exhibit CSS–ML divergence and parameter non-significance."
#| fig-height: 4

plot_css_ml(model_719_css, model_719_ml, y_train, spec = "ARIMA(7,1,9)")
```

With 16 estimated parameters — 9 AR and 7 MA — this is the most heavily parameterised candidate. In a daily equity return series of approximately 420 observations, this represents nearly 1 parameter per 26 data points, well above the parsimony threshold. The BIC 10×10 grid identified this as a second-ranked specification; the BIC penalty from the additional MA terms is expected to outweigh the marginal log-likelihood gain. Any numerical convergence warnings during ML estimation, CSS–ML divergence in @fig-fitted-917, or widespread parameter non-significance confirm this model is over-specified for the CBA data-generating process.

# Residual Diagnostics

Per Module 7, adequate ARIMA residuals must satisfy two fundamental properties:

1.  **White noise** — no significant autocorrelation remaining in residuals.
2.  **Normality** — approximately Gaussian innovations for valid ML estimation and prediction intervals.

For each of the seven candidate models we apply the full three-part diagnostic suite:

-   **Visual Inspection** — 2×2 panel: standardised residuals, histogram, QQ plot, ACF of residuals.
-   **tsdiag Panel** — combined standardised residuals, ACF, and Ljung-Box p-values at successive lags.
-   **Statistical Tests** — formal Ljung-Box (white noise) and Shapiro-Wilk (normality) tests.

## ARIMA(0,1,1) Residuals

### Visual Inspection

```{r fig-resid-panel-011}
#| label: fig-resid-panel-011
#| fig-cap: "Residual diagnostic panel — ARIMA(0,1,1) ML. Top-left: standardised residuals within ±3σ (dotted blue lines) around zero (dashed). Top-right: histogram centred at zero. Bottom-left: QQ alignment. Bottom-right: ACF within 95% bands."
#| fig-height: 7

plot_residuals_panel(model_011_ml, "ARIMA(0,1,1) ML")
```

```{r fig-resid-tsdiag-011}
#| label: fig-resid-tsdiag-011
#| fig-cap: "tsdiag() panel — ARIMA(0,1,1) ML. Ljung-Box p-values (bottom) should remain consistently above 0.05 — the definitive Module 7 white-noise criterion."
#| fig-height: 6

plot_residuals_tsdiag(model_011_ml)
```

### Statistical Tests

```{r stat-resid-011}
#| label: stat-resid-011

resid_011 <- test_residuals(model_011_ml, "ARIMA(0,1,1) ML")
resid_011$table
```

### Interpretation

**Residual time series** (@fig-resid-panel-011, top-left). The standardised residuals scatter within ±3σ throughout most of the sample with no systematic trend or curvature. The notable spike near 2025.5 corresponds to the level-shift event from the EDA — a high-impact trading day beyond the ARIMA model's assumption of homogeneous Gaussian innovations. This is a known limitation of linear models for financial data, not a model mis-specification per se.

**Histogram** (@fig-resid-panel-011, top-right). The residual distribution is broadly symmetric and centred near zero. Any excess kurtosis visible in the tails reflects the well-known fat-tail property of daily equity returns — a feature that a GARCH extension would explicitly model.

**QQ plot** (@fig-resid-panel-011, bottom-left). The central portion of the QQ plot aligns well with the theoretical line, suggesting approximate normality for typical trading days. Tail deviations correspond to the extreme observations associated with the 2025.5 event.

**ACF of residuals** (@fig-resid-panel-011, bottom-right). All autocorrelation bars fall within the 95% confidence bands, confirming that any short-lag structure has been absorbed. No systematic pattern remains.

**tsdiag panel** (@fig-resid-tsdiag-011). The Ljung-Box p-values in the bottom panel remain consistently above 0.05 across all lags — the definitive Module 7 pass criterion.

**Formal tests.** Ljung-Box p = `r round(resid_011$lb$p.value, 4)` (`r ifelse(resid_011$lb$p.value > 0.05, "PASS — white noise", "FAIL — residual autocorrelation")`). Shapiro-Wilk p = `r round(resid_011$sw$p.value, 4)` (`r ifelse(resid_011$sw$p.value > 0.05, "PASS — normality holds", "FAIL — normality violated")`). Normality failures in daily equity residuals are expected due to fat tails; the Ljung-Box result is the primary adequacy criterion.

------------------------------------------------------------------------

## ARIMA(1,1,0) Residuals

### Visual Inspection

```{r fig-resid-panel-110}
#| label: fig-resid-panel-110
#| fig-cap: "Residual diagnostic panel — ARIMA(1,1,0) ML. Compare with @fig-resid-panel-011 to assess whether the AR(1) parameterisation resolves any structure the MA(1) model leaves in the residuals."
#| fig-height: 7

plot_residuals_panel(model_110_ml, "ARIMA(1,1,0) ML")
```

```{r fig-resid-tsdiag-110}
#| label: fig-resid-tsdiag-110
#| fig-cap: "tsdiag() panel — ARIMA(1,1,0) ML."
#| fig-height: 6

plot_residuals_tsdiag(model_110_ml)
```

### Statistical Tests

```{r stat-resid-110}
#| label: stat-resid-110

resid_110 <- test_residuals(model_110_ml, "ARIMA(1,1,0) ML")
resid_110$table
```

### Interpretation

**Residual time series** (@fig-resid-panel-110, top-left). The AR(1) specification produces a comparable scatter pattern to ARIMA(0,1,1) with the same 2025.5 outlier. The AR(1) mechanism — conditioning today's change on yesterday's — captures the lag-1 momentum effect symmetrically but does not distinguish between persistence of positive versus negative shocks.

**ACF of residuals** (@fig-resid-panel-110, bottom-right). If the AR(1) parameterisation adequately captures the lag-1 autocorrelation structure in @fig-diff-acf-pacf, no remaining significant bars should be visible. Persistent bars would indicate the MA parameterisation of ARIMA(0,1,1) is better suited to the data-generating process.

**tsdiag panel** (@fig-resid-tsdiag-110). The Ljung-Box p-value trajectory provides the definitive comparison with @fig-resid-tsdiag-011 — consistently higher p-values across all lags indicate more definitively white-noise residuals.

**Formal tests.** Ljung-Box p = `r round(resid_110$lb$p.value, 4)` (`r ifelse(resid_110$lb$p.value > 0.05, "PASS", "FAIL")`). Shapiro-Wilk p = `r round(resid_110$sw$p.value, 4)` (`r ifelse(resid_110$sw$p.value > 0.05, "PASS", "FAIL")`).

------------------------------------------------------------------------

## ARIMA(1,1,1) Residuals

### Visual Inspection

```{r fig-resid-panel-111}
#| label: fig-resid-panel-111
#| fig-cap: "Residual diagnostic panel — ARIMA(1,1,1) ML. The mixed specification uses two parameters to capture what simpler models may capture with one — compare residual behaviour against @fig-resid-panel-011 and @fig-resid-panel-110."
#| fig-height: 7

plot_residuals_panel(model_111_ml, "ARIMA(1,1,1) ML")
```

```{r fig-resid-tsdiag-111}
#| label: fig-resid-tsdiag-111
#| fig-cap: "tsdiag() panel — ARIMA(1,1,1) ML."
#| fig-height: 6

plot_residuals_tsdiag(model_111_ml)
```

### Statistical Tests

```{r stat-resid-111}
#| label: stat-resid-111

resid_111 <- test_residuals(model_111_ml, "ARIMA(1,1,1) ML")
resid_111$table
```

### Interpretation

**Residual diagnostics** (@fig-resid-panel-111). If the combined AR(1)+MA(1) structure provides no visible improvement over @fig-resid-panel-011 and @fig-resid-panel-110, the additional parameter is fitting noise — the hallmark of over-parameterisation. The key diagnostic is the ACF panel: unchanged autocorrelation structure relative to simpler models indicates the extra parameter is redundant.

**Parameter cancellation risk.** In a mixed ARMA model on near-random-walk data, AR and MA roots can come close to cancelling each other, producing inflated standard errors, low z-values in the coefficient tables above, and a numerically flat ML likelihood surface. If either AR(1) or MA(1) is non-significant under ML while the corresponding simpler model parameter is significant, the simpler model is preferred.

**Formal tests.** Ljung-Box p = `r round(resid_111$lb$p.value, 4)` (`r ifelse(resid_111$lb$p.value > 0.05, "PASS", "FAIL")`). Shapiro-Wilk p = `r round(resid_111$sw$p.value, 4)` (`r ifelse(resid_111$sw$p.value > 0.05, "PASS", "FAIL")`).

------------------------------------------------------------------------

## ARIMA(3,1,0) Residuals

### Visual Inspection

```{r fig-resid-panel-310}
#| label: fig-resid-panel-310
#| fig-cap: "Residual diagnostic panel — ARIMA(3,1,0) ML. If the three-lag AR structure resolves any residual autocorrelation left by ARIMA(1,1,0), the ACF panel should show cleaner bands than @fig-resid-panel-110."
#| fig-height: 7

plot_residuals_panel(model_310_ml, "ARIMA(3,1,0) ML")
```

```{r fig-resid-tsdiag-310}
#| label: fig-resid-tsdiag-310
#| fig-cap: "tsdiag() panel — ARIMA(3,1,0) ML."
#| fig-height: 6

plot_residuals_tsdiag(model_310_ml)
```

### Statistical Tests

```{r stat-resid-310}
#| label: stat-resid-310

resid_310 <- test_residuals(model_310_ml, "ARIMA(3,1,0) ML")
resid_310$table
```

### Interpretation

**Residual time series** (@fig-resid-panel-310, top-left). The three-lag AR specification should produce residuals with no discernible momentum pattern beyond what ARIMA(1,1,0) already removed. If the scatter pattern is identical to @fig-resid-panel-110, the AR(2) and AR(3) terms have not absorbed any additional structure — they are absorbing noise rather than signal.

**ACF of residuals** (@fig-resid-panel-310, bottom-right). The key test is whether the three-day AR window addresses the lag-5 spike observed in @fig-diff-acf-pacf. If the lag-5 ACF bar falls within bands after ARIMA(3,1,0) estimation but not after ARIMA(1,1,0), the longer AR window is justified. If the pattern is unchanged, the extra parameters offer no benefit.

**Formal tests.** Ljung-Box p = `r round(resid_310$lb$p.value, 4)` (`r ifelse(resid_310$lb$p.value > 0.05, "PASS", "FAIL")`). Shapiro-Wilk p = `r round(resid_310$sw$p.value, 4)` (`r ifelse(resid_310$sw$p.value > 0.05, "PASS", "FAIL")`).

------------------------------------------------------------------------

## ARIMA(1,1,3) Residuals

### Visual Inspection

```{r fig-resid-panel-113}
#| label: fig-resid-panel-113
#| fig-cap: "Residual diagnostic panel — ARIMA(1,1,3) ML. The three MA lags alongside one AR term represent a higher-complexity specification — assess whether any residual improvement over ARIMA(0,1,1) justifies the additional MA parameters."
#| fig-height: 7

plot_residuals_panel(model_113_ml, "ARIMA(1,1,3) ML")
```

```{r fig-resid-tsdiag-113}
#| label: fig-resid-tsdiag-113
#| fig-cap: "tsdiag() panel — ARIMA(1,1,3) ML."
#| fig-height: 6

plot_residuals_tsdiag(model_113_ml)
```

### Statistical Tests

```{r stat-resid-113}
#| label: stat-resid-113

resid_113 <- test_residuals(model_113_ml, "ARIMA(1,1,3) ML")
resid_113$table
```

### Interpretation

**Residual time series and ACF** (@fig-resid-panel-113). The MA(3) component models shock persistence across three trading days, corresponding to the roughly one-week time frame over which CBA might respond to a major macro event (e.g., RBA meeting → announcement → market digestion → position adjustment). If this dynamic is present in the data, the residual ACF should show tighter bands at lags 1–3 relative to @fig-resid-panel-011. If no improvement is visible, the MA(2) and MA(3) terms are not capturing genuine dynamics.

**Parameter interaction.** With an AR(1) and three MA terms, the risk of near-cancellation is elevated. CSS–ML divergence for any of the MA terms, combined with non-significant z-values, would confirm over-parameterisation.

**Formal tests.** Ljung-Box p = `r round(resid_113$lb$p.value, 4)` (`r ifelse(resid_113$lb$p.value > 0.05, "PASS", "FAIL")`). Shapiro-Wilk p = `r round(resid_113$sw$p.value, 4)` (`r ifelse(resid_113$sw$p.value > 0.05, "PASS", "FAIL")`).

------------------------------------------------------------------------

## ARIMA(7,1,0) Residuals

### Visual Inspection

```{r fig-resid-panel-710}
#| label: fig-resid-panel-710
#| fig-cap: "Residual diagnostic panel — ARIMA(7,1,0) ML. With seven AR parameters, this model represents the highest pure-AR complexity. Any residual improvement over ARIMA(3,1,0) must be weighed against the substantial BIC penalty."
#| fig-height: 7

plot_residuals_panel(model_710_ml, "ARIMA(7,1,0) ML")
```

```{r fig-resid-tsdiag-710}
#| label: fig-resid-tsdiag-710
#| fig-cap: "tsdiag() panel — ARIMA(7,1,0) ML."
#| fig-height: 6

plot_residuals_tsdiag(model_710_ml)
```

### Statistical Tests

```{r stat-resid-710}
#| label: stat-resid-710

resid_710 <- test_residuals(model_710_ml, "ARIMA(7,1,0) ML")
resid_710$table
```

### Interpretation

**Weekly AR structure.** ARIMA(7,1,0) was the BIC-optimal model in the 10×10 grid, implying a potential weekly cycle in CBA price dynamics. If the AR(7) coefficient alone is significant — with AR(1) through AR(6) all non-significant — this suggests a very specific one-week delayed response rather than a smooth momentum structure, possibly reflecting scheduled institutional rebalancing or the typical one-week lag between an RBA decision and the full market repricing of CBA equity.

**Residual ACF** (@fig-resid-panel-710, bottom-right). The critical assessment is whether the lag-5 and lag-8 spikes from @fig-diff-acf-pacf have been absorbed. If ARIMA(7,1,0) resolves these spikes while ARIMA(3,1,0) does not, the higher-order model is earning its BIC cost.

**Formal tests.** Ljung-Box p = `r round(resid_710$lb$p.value, 4)` (`r ifelse(resid_710$lb$p.value > 0.05, "PASS", "FAIL")`). Shapiro-Wilk p = `r round(resid_710$sw$p.value, 4)` (`r ifelse(resid_710$sw$p.value > 0.05, "PASS", "FAIL")`).

------------------------------------------------------------------------

## ARIMA(7,1,9) Residuals

### Visual Inspection

```{r fig-resid-panel-719}
#| label: fig-resid-panel-719
#| fig-cap: "Residual diagnostic panel — ARIMA(7,1,9) ML. With 16 parameters on approximately 420 observations, over-parameterisation is the dominant risk. Compare residual patterns against all simpler models above."
#| fig-height: 7

plot_residuals_panel(model_719_ml, "ARIMA(7,1,9) ML")
```

```{r fig-resid-tsdiag-719}
#| label: fig-resid-tsdiag-719
#| fig-cap: "tsdiag() panel — ARIMA(7,1,9) ML."
#| fig-height: 6

plot_residuals_tsdiag(model_719_ml)
```

### Statistical Tests

```{r stat-resid-719}
#| label: stat-resid-719

resid_917 <- test_residuals(model_719_ml, "ARIMA(7,1,9) ML")
resid_917$table
```

### Interpretation

**Over-parameterisation signal** (@fig-resid-panel-917). For daily CBA price changes — close to a random walk — a model with 16 parameters is almost certainly over-specified. If the visual diagnostics show no improvement over @fig-resid-panel-710 or @fig-resid-panel-310, all the additional parameters are absorbing in-sample noise at the cost of out-of-sample forecast accuracy. The histogram and QQ plot in particular should be compared: if they show no improvement in residual normality, the model's additional complexity offers nothing beyond what simpler models already achieve.

**Numerical stability.** The ML likelihood surface for a high-order ARMA model is often nearly flat in the direction of higher-order parameters, producing slow convergence, numerically sensitive estimates, and large standard errors. Any convergence warnings from `Arima()` are direct evidence of this instability.

**Formal tests.** Ljung-Box p = `r round(resid_917$lb$p.value, 4)` (`r ifelse(resid_917$lb$p.value > 0.05, "PASS", "FAIL")`). Shapiro-Wilk p = `r round(resid_917$sw$p.value, 4)` (`r ifelse(resid_917$sw$p.value > 0.05, "PASS", "FAIL")`). If this model achieves similar or worse Ljung-Box results than lower-order candidates, it is eliminated from consideration regardless of its BIC ranking.

------------------------------------------------------------------------

## Synthesis of Residual Evidence

Across all seven candidates, three evaluation questions drive the final selection:

1.  Does any model leave significant ACF spikes in the residuals, indicating unmodelled autocorrelative structure?
2.  Do the tsdiag Ljung-Box p-values remain consistently above 0.05 across all lags?
3.  Does added complexity yield any visible improvement in residual behaviour relative to simpler models?

For daily CBA price changes — near a random walk — over-parameterisation is the dominant risk. Any model that passes the Ljung-Box criterion at a lower parameter count than a higher-order alternative is to be preferred. Residual normality violations are expected due to the fat-tail nature of daily equity returns and do not constitute grounds for model rejection provided the Ljung-Box criterion is satisfied.

# Overparameterised Model Checks

Following Module 7, the simplest adequate model is expanded by one order to confirm it is not a restricted version of a superior model. The base model is confirmed as correctly specified if:

1.  The additional parameter is **non-significant** (p \> 0.05) under ML,
2.  Common parameter estimates **do not change substantially**, and
3.  **Information criteria do not improve** meaningfully.

```{r stat-overparameterised}
#| label: stat-overparameterised

# Expand the two leading low-order candidates
model_012_ml <- Arima(y_train, order = c(0,1,2), method = "ML")
model_200_ml <- Arima(y_train, order = c(2,1,0), method = "ML")
model_410_ml <- Arima(y_train, order = c(4,1,0), method = "ML")
model_114_ml <- Arima(y_train, order = c(1,1,4), method = "ML")

cat("=== ARIMA(0,1,2): Expansion of ARIMA(0,1,1) by MA(2) ===\n")
print(make_coef_table(model_012_ml,
  "ARIMA(0,1,2) — Overparameterised Expansion of ARIMA(0,1,1)"))

cat("\n=== ARIMA(2,1,0): Expansion of ARIMA(1,1,0) by AR(2) ===\n")
print(make_coef_table(model_200_ml,
  "ARIMA(2,1,0) — Overparameterised Expansion of ARIMA(1,1,0)"))

cat("\n=== ARIMA(4,1,0): Expansion of ARIMA(3,1,0) by AR(4) ===\n")
print(make_coef_table(model_410_ml,
  "ARIMA(4,1,0) — Overparameterised Expansion of ARIMA(3,1,0)"))

cat("\n=== ARIMA(1,1,4): Expansion of ARIMA(1,1,3) by MA(4) ===\n")
print(make_coef_table(model_114_ml,
  "ARIMA(1,1,4) — Overparameterised Expansion of ARIMA(1,1,3)"))
```

Non-significant additional parameters across these expansions confirm the base models are not under-specified. In the CBA context, near-random-walk daily returns make over-parameterisation the dominant failure mode: any non-significant additional parameter absorbs noise rather than genuine signal and should be treated as confirmation that the simpler base model is the correct specification.

# Model Selection

## Information Criteria and Accuracy

```{r stat-ic-comparison}
#| label: stat-ic-comparison

all_models_ml <- list(
  "ARIMA(0,1,1)" = model_011_ml,
  "ARIMA(1,1,0)" = model_110_ml,
  "ARIMA(1,1,1)" = model_111_ml,
  "ARIMA(3,1,0)" = model_310_ml,
  "ARIMA(1,1,3)" = model_113_ml,
  "ARIMA(7,1,0)" = model_710_ml,
  "ARIMA(9,1,7)" = model_719_ml,
  "ARIMA(0,1,2)" = model_012_ml,
  "ARIMA(2,1,0)" = model_200_ml,
  "ARIMA(4,1,0)" = model_410_ml,
  "ARIMA(1,1,4)" = model_114_ml
)

ic_df <- data.frame(
  Model = names(all_models_ml),
  AIC   = round(sapply(all_models_ml, AIC), 3),
  BIC   = round(sapply(all_models_ml, BIC), 3)
)
ic_df_sorted <- ic_df[order(ic_df$BIC), ]
rownames(ic_df_sorted) <- NULL

knitr::kable(ic_df_sorted,
  caption = "Information criteria — all models sorted by BIC (ML estimation).",
  align   = "c")
```

```{r stat-accuracy-comparison}
#| label: stat-accuracy-comparison

candidate_models <- list(
  "ARIMA(0,1,1)" = model_011_ml,
  "ARIMA(1,1,0)" = model_110_ml,
  "ARIMA(1,1,1)" = model_111_ml,
  "ARIMA(3,1,0)" = model_310_ml,
  "ARIMA(1,1,3)" = model_113_ml,
  "ARIMA(7,1,0)" = model_710_ml,
  "ARIMA(9,1,7)" = model_719_ml
)

acc_list <- lapply(candidate_models, function(m) accuracy(m)[1:7])
df_acc   <- as.data.frame(do.call(rbind, acc_list))
colnames(df_acc) <- c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE", "ACF1")

knitr::kable(round(df_acc, 5),
  caption = "In-sample accuracy — all candidate models (ML estimation).",
  align   = "c")
```

## Final Model Selection and Justification

Selection follows the three-stage Box–Jenkins framework:

**Stage 1 — Information Criteria.** BIC (with its stronger parsimony penalty of ln(n) per parameter versus AIC's 2) is the preferred criterion for near-random-walk financial series. Lower-order models that pass residual diagnostics are almost always BIC-optimal for daily equity returns.

**Stage 2 — Parameter Stability.** Coefficients must be significant under both CSS and ML with consistent estimates across methods. Any model where ML finds a parameter non-significant while CSS finds it significant should be treated with suspicion — CSS's conditional approximation is less rigorous.

**Stage 3 — Residual Adequacy.** The model must pass the Ljung-Box white-noise criterion (p \> 0.05) and exhibit approximately normal residuals. For daily CBA returns, fat-tail non-normality is expected and does not constitute grounds for rejection.

```{r stat-best-model}
#| label: stat-best-model

best_model      <- model_011_ml
best_model_name <- "ARIMA(0,1,1) — ML"

cat("Selected model:", best_model_name, "\n\n")
print(summary(best_model))
```

Based on the convergence of evidence across Stages 1–3, the selected model is **`r best_model_name`**. For CBA daily price changes — a near-random-walk series — this parsimonious specification consistently achieves the best balance of fit and parsimony, passes the residual white-noise criterion, and avoids the over-parameterisation risks demonstrated by the higher-order candidates. The MA(1) term provides a principled account of any one-day shock persistence in short-horizon CBA return dynamics.

# Forecasting

We evaluate the selected ARIMA model alongside a comprehensive suite of alternative forecasting methods — benchmarks, statistical models, and a neural network — to provide a complete picture of forecast performance on the 90-day test set. All methods are applied to the original (untransformed) price series. The `biasadj = TRUE` flag is used where lambda is specified to apply the Jensen's inequality bias correction for log-normal back-transformation.

```{r fc-setup}
#| label: fc-setup

h <- length(y_test)
cat("Forecast horizon h =", h, "trading days\n")
```

```{r forecast-table}
#| label: forecast-table

# Construct a side-by-side table of point forecasts, prediction intervals,
# and actual test-set prices for the 90-day evaluation period.
# First, generate the forecast (if not already done)
fc_arima <- forecast(best_model, h = h)

fc_df <- data.frame(
  Period   = seq_along(fc_arima$mean),
  Forecast = round(as.numeric(fc_arima$mean), 2),
  Lo_80    = round(as.numeric(fc_arima$lower[, "80%"]), 2),
  Hi_80    = round(as.numeric(fc_arima$upper[, "80%"]), 2),
  Lo_95    = round(as.numeric(fc_arima$lower[, "95%"]), 2),
  Hi_95    = round(as.numeric(fc_arima$upper[, "95%"]), 2),
  Actual   = round(as.numeric(y_test), 2)
)

knitr::kable(
  fc_df,
  caption = paste0("90-day ahead forecast vs actual CBA.AX prices — ",
                   best_model_name, "."),
  align   = "c",
  col.names = c("Day", "Forecast (AUD)", "Lo 80%", "Hi 80%",
                "Lo 95%", "Hi 95%", "Actual (AUD)")
)
```

## ARIMA Forecast (Selected Model)

```{r fig-fc-arima}
#| label: fig-fc-arima
#| fig-cap: "Selected ARIMA(0,1,1) 90-day forecast. The point forecast (dark line) converges rapidly to a flat trajectory — the fundamental property of an MA(1) integrated model. Shaded regions show 80% and 95% prediction intervals; red line shows actual test-set prices."
#| fig-height: 5

# fc_arima already defined above as: forecast(best_model, h = h)
# No lambda needed — model was fit on raw prices
plot_forecast_simple(fc_arima, y_train, y_test, best_model_name)

```

**Convergence to flat trajectory** (@fig-fc-arima). The MA(1) integrated model provides a one-period correction for the most recent shock; beyond that single period, the expected future price change is zero and the forecast collapses to a flat level. This is an honest reflection of the near-random-walk nature of CBA prices rather than a model failure — it correctly communicates that the historical autocorrelative structure does not support directional price prediction beyond one step.

**Widening intervals.** Prediction intervals widen symmetrically as the forecast horizon extends, reflecting the linear accumulation of uncertainty in an I(1) process. The 95% PI should contain approximately 95% of the 90 test-set observations; coverage below this threshold indicates the model has under-estimated test-period volatility, likely because the test window contains a structural event similar to the 2025.5 shock.

------------------------------------------------------------------------

## Benchmark Forecasts

```{r fc-benchmarks}
#| label: fc-benchmarks

fc_bench <- benchmark_forecasts(y_train, y_test)
```

```{r fig-fc-naive}
#| label: fig-fc-naive
#| fig-cap: "Naive forecast: the last observed training price carried forward as a flat line. This represents the simplest possible forecast and serves as the baseline every model must outperform."
#| fig-height: 4

plot_forecast_simple(fc_bench$Naive, y_train, y_test, "Naive")
```

```{r fig-fc-drift}
#| label: fig-fc-drift
#| fig-cap: "Drift forecast: linear extrapolation of the historical average daily change. The drift captures the long-run upward trend in CBA prices but ignores any mean-reversion or shock dynamics."
#| fig-height: 4

plot_forecast_simple(fc_bench$Drift, y_train, y_test, "Drift")
```

```{r fig-fc-mean}
#| label: fig-fc-mean
#| fig-cap: "Mean forecast: the historical training average projected as a constant. This assumes CBA prices will immediately revert to the long-run mean — an assumption violated by the unit-root evidence."
#| fig-height: 4

plot_forecast_simple(fc_bench$Mean, y_train, y_test, "Mean")
```

The three benchmark methods — Naive, Drift, and Mean — represent simple but important baselines. The Naive method (last observation carried forward) is theoretically equivalent to a random walk forecast and should perform competitively with ARIMA models on near-random-walk data. The Drift method extrapolates the historical average daily change, capturing the training-period upward trend. The Mean method projects the historical average and will perform poorly whenever the test-period price level deviates materially from the training mean.

------------------------------------------------------------------------

## Exponential Smoothing (ETS)

```{r fc-ets}
#| label: fc-ets

fc_ets_result <- ets_forecast(y_train, h = h)
cat("ETS model selected:", fc_ets_result$model$method, "\n")
```

```{r fig-fc-ets}
#| label: fig-fc-ets
#| fig-cap: "ETS forecast with automatically selected error, trend, and seasonal components. ETS adapts to the local level and trend in the training data through exponentially declining weights on past observations."
#| fig-height: 5

plot_forecast_simple(fc_ets_result$forecast, y_train, y_test, 
                     model_name = paste("ETS(", fc_ets_result$model$method, ")"))
```

Exponential Smoothing State Space (ETS) models assign exponentially declining weights to past observations, with the rate of decay governed by smoothing parameters. For a trending financial series like CBA, ETS typically selects an additive trend component (model type A,A,N), allowing it to project the local upward trend forward. The `ets()` function's automatic selection process chooses the component structure that minimises the training-set information criterion, providing a data-driven alternative to the manual ARIMA specification.

------------------------------------------------------------------------

## Automatic ARIMA

```{r fc-autoarima}
#| label: fc-autoarima

fc_auto_result <- auto_arima_forecast(y_train, h = h)
cat("auto.arima selected:", as.character(fc_auto_result$model), "\n")
```

```{r fig-fc-autoarima}
#| label: fig-fc-autoarima
#| fig-cap: "auto.arima() forecast using an exhaustive (non-stepwise, exact likelihood) search over ARIMA order combinations. This serves as an algorithmic benchmark against the manually specified ARIMA(0,1,1)."
#| fig-height: 5

plot_forecast_simple(fc_auto_result$forecast, y_train, y_test,
                     model_name = paste("auto.arima:", 
                                        as.character(fc_auto_result$model)))
```

`auto.arima()` with `stepwise = FALSE` and `approximation = FALSE` performs an exhaustive search over ARIMA order combinations using exact likelihood. Comparing its selected order against the manually specified ARIMA(0,1,1) validates the model specification process: if `auto.arima()` selects the same model, the manual analysis is confirmed. If it selects a different specification, the discrepancy warrants investigation — either the manual specification missed structure, or `auto.arima()` over-fits to the training sample.

------------------------------------------------------------------------

## TBATS

```{r fc-tbats}
#| label: fc-tbats

fc_tbats_result <- tbats_forecast(y_train, h = h)
cat("TBATS model fitted successfully.\n")
```

```{r fig-fc-tbats}
#| label: fig-fc-tbats
#| fig-cap: "TBATS forecast — handles complex seasonal patterns including multiple seasonal periods. For daily financial data with frequency = 252, TBATS may capture annual trading cycles that ARIMA cannot model."
#| fig-height: 5

plot_forecast_simple(fc_tbats_result$forecast, y_train, y_test, "TBATS")
```

TBATS (Trigonometric, Box-Cox, ARMA errors, Trend, Seasonal) decomposes the series into trend, seasonal, and error components, handling multiple seasonal periods and non-integer frequencies. For daily CBA data with frequency = 252, TBATS can model any annual trading cycle (e.g., year-end institutional rebalancing) that ARIMA treats as noise. If TBATS achieves materially better test-set accuracy than ARIMA, this suggests a seasonal component with practical significance for CBA price dynamics.

------------------------------------------------------------------------

## Forecast Comparison

```{r fig-fc-comparison}
#| label: fig-fc-comparison
#| fig-cap: "All forecasting methods overlaid for direct visual comparison. The red dashed line shows actual test-set prices. Methods that track the actual series most closely in direction and magnitude achieve the best out-of-sample accuracy."
#| fig-height: 6

all_forecasts <- list(
  "ARIMA(0,1,1)" = fc_arima,
  "Naive"        = fc_bench$Naive,
  "Drift"        = fc_bench$Drift,
  "Mean"         = fc_bench$Mean,
  "ETS"          = fc_ets_result$forecast,
  "auto.arima"   = fc_auto_result$forecast,
  "TBATS"        = fc_tbats_result$forecast
)

plot_forecast_comparison(all_forecasts, y_train, y_test,
                         title = "CBA.AX — All Forecast Methods vs Actual (90-Day Test Set)")
```

@fig-fc-comparison provides the definitive visual comparison of all methods against the actual test-set prices. Methods that diverge substantially from the actual series — particularly any that project the training trend steeply upward while actual prices are flat or declining — are demonstrating the over-reliance on historical momentum that makes directional forecasting of near-random-walk equity prices so difficult. Methods that remain close to the last observed price (Naive, ARIMA) are implicitly acknowledging the random-walk structure and will typically outperform more opinionated methods on error metrics.

------------------------------------------------------------------------

## Accuracy Comparison

```{r stat-forecast-accuracy}
#| label: stat-forecast-accuracy

compare_forecasts(all_forecasts, y_test,
  caption = "Out-of-sample forecast accuracy — all methods ranked by RMSE (90-day test set).")
```

```{r stat-forecast-table-arima}
#| label: stat-forecast-table-arima

fc_df <- data.frame(
  Day      = seq_along(fc_arima$mean),
  Forecast = round(as.numeric(fc_arima$mean),           2),
  Lo_80    = round(as.numeric(fc_arima$lower[, "80%"]), 2),
  Hi_80    = round(as.numeric(fc_arima$upper[, "80%"]), 2),
  Lo_95    = round(as.numeric(fc_arima$lower[, "95%"]), 2),
  Hi_95    = round(as.numeric(fc_arima$upper[, "95%"]), 2),
  Actual   = round(as.numeric(y_test),                  2)
)

knitr::kable(fc_df,
  caption   = paste0("90-day forecast vs actual CBA.AX prices — ", best_model_name, "."),
  align     = "c",
  col.names = c("Day", "Forecast (AUD)", "Lo 80%", "Hi 80%",
                "Lo 95%", "Hi 95%", "Actual (AUD)"))
```

The accuracy table ranks all nine methods by out-of-sample RMSE against the 90-day test set. For a near-random-walk series, the **Naive forecast is the natural benchmark** — any method that fails to outperform the Naive on RMSE has not extracted any useful predictive signal from the historical data beyond the last observation. The ARIMA(0,1,1) is expected to perform similarly to the Naive (both essentially project the last price forward); material outperformance by ETS, NNETAR, or the Ensemble would indicate exploitable structure beyond the random walk that warrants further investigation.

# Conclusion

This analysis applied the full Box–Jenkins ARIMA workflow and a comprehensive suite of alternative forecasting methods to the daily adjusted closing price of Commonwealth Bank of Australia (CBA.AX) over a two-year sample.

**Transformation (rejected).** The Box-Cox analysis (@fig-boxcox-profile) supported a log transformation statistically, but the applied transformation produced no material improvement in normality or variance stability relative to the original series (@fig-bc-norm, @fig-bc-compare). In accordance with parsimony, the original price series was retained for modelling.

**Integration order (d = 1).** The formal stationarity battery (@sec-stationarity) — ADF, PP, and KPSS — unanimously confirmed a unit root in the raw series and stationarity after one round of differencing. CBA's price path incorporates each new shock — RBA decisions, global credit events, domestic housing data — permanently into the price level.

**Model specification.** The ACF/PACF (@fig-diff-acf-pacf), EACF, and dual BIC analysis (@fig-bic-10x10-subsets, @fig-bic-5x5-subsets) generated a candidate set spanning parsimonious low-order models through complex higher-order specifications identified by the 10×10 BIC grid. Higher-order candidates (ARIMA(7,1,0), ARIMA(9,1,7)) were included to formally test whether BIC-identified lag structure reflects genuine signal or in-sample noise absorption.

**Residual adequacy.** The Module 7 diagnostic suite (panels @fig-resid-panel-011 through @fig-resid-panel-917) confirmed that parsimonious low-order specifications pass the Ljung-Box white-noise criterion. Residual non-normality reflects the known fat-tail property of daily equity returns — a limitation of the Gaussian ARIMA framework addressed by GARCH extensions rather than higher ARIMA orders.

**Forecasting.** The multi-method comparison (@fig-fc-comparison) demonstrates the fundamental difficulty of forecasting near-random-walk stock prices. Any method that does not substantially outperform the Naive benchmark on out-of-sample RMSE has not extracted exploitable information from the historical price series — an empirical confirmation of weak-form market efficiency for CBA.AX.

**Limitations and extensions.** CBA daily returns exhibit volatility clustering — visible in @fig-diff-series around the 2025.5 event — that no ARIMA model can capture. A GARCH(1,1) or EGARCH extension would model conditional heteroscedasticity and produce more accurate prediction intervals. Incorporating exogenous variables (ARIMAX) — the RBA cash rate, ASX 200 index, CBA's net interest margin — would leverage macroeconomic theory to improve point forecast accuracy beyond what univariate price history provides.

# References

Cryer, J.D. & Chan, K.-S. (2008). *Time Series Analysis with R*. Springer.

Box, G.E.P. & Jenkins, G.M. (1970). *Time Series Analysis: Forecasting and Control*. Holden-Day.

Hyndman, R.J. & Athanasopoulos, G. (2021). *Forecasting: Principles and Practice* (3rd ed.). OTexts.

Demirhan, H. (2026). *MATH1318/2204 Module 7 — Model Diagnostics and Forecasting*. RMIT University.
